\documentclass{article}
% Use \usepackage[final]{neurips_2026} for camera-ready
\usepackage[preprint]{neurips_2026}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\usepackage{url}
\usepackage{booktabs}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{nicefrac}
\usepackage{microtype}
\usepackage{xcolor}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{pgfplots}
\usepackage{tikz}
\usetikzlibrary{positioning, arrows.meta, shapes.geometric, fit, calc}
\usepackage{pifont}
\usepackage{amsthm}
\newcommand{\cmark}{\ding{51}}
\newcommand{\xmark}{\ding{55}}
\newtheorem{definition}{Definition}
\newtheorem{proposition}{Proposition}
\newcommand{\nonterminal}[1]{\texttt{#1}}
\pgfplotsset{compat=1.18}

\title{Raw-Routed Adapters and Architecture Search for Time Series Foundation Models}

\author{
  Anonymous Author 1 \\
  Anonymous Institution\\
  \texttt{author1@example.com}
  \And
  Anonymous Author 2 \\
  Anonymous Institution\\
  \texttt{author2@example.com}
  \And
  Anonymous Author 3 \\
  Anonymous Institution\\
  \texttt{author3@example.com}
}

\begin{document}

\maketitle

\begin{abstract}
Adapting pretrained time series foundation models (TSFMs) to downstream tasks currently relies on static, single-head adapters. While dynamically routing across diverse adapter topologies via mixture-of-experts (MoE) offers a path to task-specific specialization, we identify a fatal structural bottleneck in temporal adaptation: \textbf{gradient co-adaptation}. We demonstrate that standard MoE routing on TSFM hidden states collapses to a single expert---routing entropy dropping to $0.000\pm0.000$ (3 seeds)---particularly when backbone layers are unfrozen, as deep representations co-adapt to reinforce the dominant expert.
To break this collapse, we propose \textbf{Raw-Routed Mixture of Adapters (RR-MoA)}. RR-MoA bypasses the backbone's internal Reversible Instance Normalization (RevIN~\citep{kim2021revin}) entirely, utilizing a lightweight gating network that reads the raw, \emph{pre-RevIN} input signal to preserve the temporal statistics (trend, amplitude, volatility) that the backbone would otherwise strip away before routing. A direct ablation shows that applying per-window RevIN to the router's input degrades MSE by $60$--$88\%$ across all three datasets, empirically isolating rawness from bypass as the source of the gains. Crucially, we uncover a \textbf{``Frozen Paradox''}: keeping the backbone strictly frozen prevents co-adaptation and yields \emph{superior} MoE routing compared to standard partial fine-tuning. Evaluated rigorously across three seeds, strictly frozen RR-MoA achieves \textbf{27/27 wins} against single-adapter baselines and \textbf{9/9 wins} against a modern PEFT baseline (LoRA~\citep{hu2022lora}) under the identical protocol, yielding a $43.5\%$ MSE reduction on ETTh1 ($0.690{\pm}0.02$), $51.1\%$ on ETTm1 ($0.572{\pm}0.07$), and $44.6\%$ on Weather ($0.289{\pm}0.01$). To populate the expert pool we further introduce \textbf{Adapter Architecture Search (AAS)}: an LLM-guided code-level search that imports cross-domain motifs (depthwise convolutions, BatchNorm gating, feature attention) into the TSFM adapter vocabulary. Finally, enforcing \textbf{Top-2 sparse routing} bounds the active expert FLOPs to $40\%$ of a dense mixture while preserving the MSE gains, establishing a compute-fair, Pareto-optimal paradigm for parameter-efficient TSFM adaptation.
\end{abstract}

\section{Introduction}
\label{sec:intro}

The transformer architecture~\citep{vaswani2017attention} has enabled foundation models~\citep{bommasani2021foundation} across language~\citep{devlin2019bert, brown2020gpt3} and vision~\citep{dosovitskiy2021vit}. Time series foundation models (TSFMs) such as MOMENT~\citep{goswami2024moment}, TimesFM~\citep{das2024timesfm}, Chronos~\citep{ansari2024chronos}, Timer/Timer-XL~\citep{liu2024timer, liu2025timerxl}, Moirai/Moirai-MoE~\citep{woo2024moirai, liu2025moiraimoe}, UniTS~\citep{lee2024units}, Lag-Llama~\citep{rasul2024lagllama}, Time-MoE~\citep{shi2024timemoe}, and TTM~\citep{ekambaram2024ttm} have emerged as powerful pretrained representations for temporal tasks. Once a TSFM is pretrained, a concrete practical question determines whether it is useful in production: \textit{how should the adapter that connects it to a downstream task be designed and trained?}

\textbf{The one-size-fits-all adapter status quo.} Current TSFM adaptation almost always reduces to the same computational template: \emph{pool the backbone's hidden states $\to$ project to the output horizon}, with a handful of standard variants---linear probing~\citep{goswami2024moment, das2024timesfm}, MLP heads~\citep{nie2023patchtst, wu2023timesnet}, LoRA~\citep{hu2022lora}, and bottleneck adapters~\citep{houlsby2019adapters}. Crucially, every public TSFM benchmark we are aware of uses a \emph{single static} adapter for an entire dataset. Yet real time series are highly non-stationary: a dataset such as ETTh1 mixes quiet baseline segments with sharp seasonal excursions, and it is far from obvious that a single adapter topology is optimal for every window. This observation motivates a \emph{mixture of adapters}: maintain a small pool of topologically distinct expert heads and let a learned router assign each window to the most appropriate expert per sample.

\textbf{Why frozen at all?} For a single dataset at a single horizon, lightweight supervised baselines like DLinear~\citep{zeng2023dlinear} obtain materially lower absolute MSE than any frozen-backbone adapter. We measure this directly on our exact normalized evaluation scale: a $49$K-parameter DLinear trained from scratch for 15 epochs reaches $0.417{\pm}0.002$ on ETTh1, $0.322{\pm}0.004$ on ETTm1, and $0.208{\pm}0.003$ on Weather (3 seeds)---roughly $40$--$80\%$ lower than our best frozen-backbone RR-MoA numbers in Table~\ref{tab:rrmoa}. Full fine-tuned specialists such as PatchTST~\citep{nie2023patchtst} would further widen this gap. \textit{Our contribution is emphatically not to replace DLinear or PatchTST in the single-dataset setting.} The frozen-backbone paradigm we target is instead motivated by three deployment regimes in which full fine-tuning is infeasible and in which re-training a DLinear per use-case is operationally expensive: (i)~\textbf{multi-tenant cloud serving}, where a single TSFM sits in GPU VRAM and serves thousands of tenants via lightweight, hot-swappable per-tenant adapters in host RAM---retraining a separate DLinear per tenant does not amortize the backbone cost and forfeits any zero-shot transfer the TSFM provides; (ii)~\textbf{on-device and edge deployment}, where one backbone is shipped once with the device image and per-site adapters are updated over low-bandwidth links---DLinear per site requires collecting each site's training data at a central location, which is often impossible for privacy or latency reasons; and (iii)~\textbf{cross-task reuse}, where the same frozen backbone drives forecasting, imputation, and (in principle) classification via different expert pools---we demonstrate the first two of these explicitly in \S\ref{sec:rrmoa} and \S4.5. Within this paradigm, the question is \emph{not} whether the frozen TSFM beats a specialist trained from scratch on the same dataset (it does not), but which adapter / routing strategy minimises the downstream error for the fixed backbone that practitioners are committed to. The regime is also \emph{future-proof}: as TSFMs scale toward billion-parameter regimes---Time-MoE~\citep{shi2024timemoe} is already a billion-parameter TSFM, and larger ones are on the horizon---full per-dataset fine-tuning becomes computationally prohibitive and increasingly prone to catastrophic forgetting of the pretrained distribution. Frozen-backbone PEFT is therefore not a concession to today's hardware; it is the only viable per-dataset adaptation path for the TSFM scale curve ahead, and our work optimises the best routing strategy within that near-future paradigm. That is the question we answer.

\textbf{The natural idea that silently fails: MoE on TSFMs.} The standard recipe for instance-level mixtures is by now well understood in NLP and vision~\citep{wang2022adamix, fedus2022switch}: train $K$ experts jointly with a gate that reads the backbone's hidden states, add a load-balancing loss, and trust the optimizer to distribute the load. When we apply this recipe to a TSFM with exactly the same five small expert heads we use throughout this paper, it does not merely perform worse than a single adapter---it \emph{collapses catastrophically}. Across 3 seeds, 3 datasets, and every setting where at least one encoder block is unfrozen, the routing distribution becomes a one-hot vector and the Shannon entropy over the 5 experts is exactly $0.000\pm0.000$ (Table~\ref{tab:adamix}): all weight is permanently concentrated on a single expert, regardless of the input.

\textbf{Diagnosis: gradient co-adaptation, not LayerNorm geometry.} This collapse is not a hyperparameter issue, nor is it explained by the generic ``LayerNorm washes out heterogeneity'' story that has been invoked for vision MoEs. It has a specific causal structure that is unique to jointly fine-tuning an MoE together with a pretrained foundation model: whichever expert the router marginally prefers in the first few gradient steps receives larger error signals; those error signals flow back through the expert into the unfrozen top of the backbone; the backbone's hidden states $\mathbf{H}$ are then jointly updated to \emph{serve that expert better}; this in turn sharpens the router's preference for the same expert on the next step. The feedback loop is positive, fast, and irreversible. In Table~\ref{tab:adamix} we show that \emph{freezing the backbone is sufficient to break this loop}: AdaMix entropy recovers from $0.000$ to $0.5$--$0.6$ when $\theta$ is held fixed. But hidden-state routing under a frozen backbone still underperforms a single fixed adapter, because the backbone's internal RevIN~\citep{kim2021revin} has already stripped the temporal statistics (trend, amplitude, volatility) that would let the router distinguish between samples. We validate this specific mechanism in \S\ref{sec:rrmoa}, Table~\ref{tab:router_input}: applying the same per-window RevIN transform to the router's input signal degrades MSE by $60$--$88\%$ across all three datasets, whereas the raw router retains its full routing gain. \emph{Why doesn't this happen in NLP or vision MoEs?} NLP tokens~\citep{fedus2022switch} and ViT patches~\citep{dosovitskiy2021vit} carry discrete semantic or spatial identities that survive LayerNorm: the feature-dimension normalization preserves per-token class-specific directions, so the router still sees well-separated hidden states. Time-series windows are characterized instead by macro-statistics---trend, amplitude, and volatility---that RevIN (and any external per-channel standardization) explicitly removes before the backbone encoder ever sees them. RR-MoA's advantage is therefore not a LayerNorm argument at all; it is specifically about restoring the router's access to the macro-statistics that TSFMs are designed to strip.

\textbf{Our fix: route on the raw pre-normalization signal, and freeze $\theta$ completely.} We propose \textbf{Raw-Routed Mixture of Adapters (RR-MoA)}, which severs gradient co-adaptation in two complementary ways. First, the router never sees $f_\theta(\mathbf{X})$; instead it reads the \emph{raw, pre-normalization} input $\mathbf{X}_\text{raw}$, which retains per-window trend, amplitude, and volatility (Table~\ref{tab:router_input} confirms that stripping these statistics via per-window RevIN alone is sufficient to erase $60$--$88\%$ of RR-MoA's advantage, even when the hidden-state bypass itself is preserved). Second, the entire TSFM backbone is kept \emph{strictly frozen} during training---every single parameter in $\theta$ has $\texttt{requires\_grad}{=}\texttt{False}$. The router and the experts are the only trainable modules. This combination reduces the inner loop of MoE training from a joint optimization over $\theta_\text{active}\cup\phi\cup\psi$ to a simpler problem in $\phi\cup\psi$ alone, and the router's input signal is a statistic the backbone cannot overwrite.

\textbf{The Frozen Paradox.} Because freezing the backbone is commonly treated as a concession to compute budgets, we expected partial unfreezing (updating the last 2 or 4 encoder blocks) to still be a useful regularization knob. It is not. In our 3\,datasets\,$\times$\,3\,freeze levels\,$\times$\,3\,seeds ablation (Table~\ref{tab:rrmoa}), strictly frozen RR-MoA is \emph{at least as good as} partially unfrozen RR-MoA on every dataset, and is strictly better on ETTh1 and ETTm1 where it delivers $43.5\%$ and $51.1\%$ MSE reductions respectively. Unfreezing more of the backbone does not buy expressivity---it buys routing collapse. We call this empirical pattern the \emph{Frozen Paradox}, and it reverses the standard NLP PEFT heuristic of ``unfreeze the top few layers for better performance.''

\textbf{Populating the expert pool with AAS.} For RR-MoA to work, the expert pool must be diverse: five copies of the same pooling head give the router nothing to route over. In the second half of the paper we formalize \textbf{Adapter Architecture Search (AAS)} as an LLM-guided code-level search over arbitrary PyTorch \texttt{nn.Module} programs. AAS acts as a cheap \emph{search-space widener} that recovers cross-domain motifs---depthwise convolutions~\citep{howard2017mobilenets}, BatchNorm gating~\citep{he2016resnet, ioffe2015batchnorm}, feature attention~\citep{hu2018senet}---from the vision and mobile architecture literatures, motifs that do not appear in any published TSFM adapter. We distill the search logs into the five canonical expert topologies used in all RR-MoA experiments so that any improvement is attributable to routing, not to extra expert capacity. We are careful to frame AAS as \emph{recovering and transferring} established motifs into the TSFM adapter vocabulary, not as discovering new algorithms.

Our contributions:

\begin{enumerate}
\item \textbf{The gradient co-adaptation diagnosis.} We identify a fundamental incompatibility between standard MoE routing and pretrained TSFMs that goes beyond prior LayerNorm intuitions: under joint adapter\,+\,backbone optimization, deep representations rapidly co-adapt with whichever expert the router favors early, starving all other experts. Empirically, AdaMix~\citep{wang2022adamix}-style hidden-state routing collapses to routing entropy $0.000\pm0.000$ across 3 seeds on every dataset once even the last 2 encoder blocks are unfrozen (\S\ref{sec:rrmoa}, Table~\ref{tab:adamix}). This is \emph{specific} to frozen/partially frozen TSFMs.

\item \textbf{Raw-Routed Mixture of Adapters (RR-MoA).} We propose routing on the raw, pre-normalization input signal rather than on the backbone's hidden states. A lightweight convolutional gate reads the raw time series to select among a fixed pool of lightweight experts that process the backbone's deep embeddings. RR-MoA achieves \textbf{27/27 wins} against single-adapter baselines across 3 datasets, 3 freeze-levels, and 3 seeds, with up to $51.1\%$ MSE reduction on ETTm1 (\S\ref{sec:rrmoa}, \S\ref{sec:main_results}). A controlled ablation that applies per-window RevIN to the router's input degrades MSE by $60$--$88\%$ across all three datasets, empirically isolating the rawness of the router signal---not merely the bypass of hidden states---as the causal mechanism (Table~\ref{tab:router_input}).

\item \textbf{The Frozen Paradox.} We show that keeping the backbone \emph{strictly frozen}---doing zero backbone parameter updates---is not a concession to efficiency but is \emph{Pareto-optimal} for routing-based adaptation. Unfreezing 2--4 encoder blocks improves single-adapter MSE only marginally while completely destroying routing diversity. Strictly frozen RR-MoA beats last-4-unfrozen RR-MoA on ETTh1 (\S\ref{sec:freeze_ablation}).

\item \textbf{Compute-fair Top-$k$ sparse routing.} To avoid conflating routing efficacy with brute-force parameter scaling, we enforce Top-2 sparse routing: only the two experts with the highest raw-signal routing probability execute per sample, bounding active expert FLOPs to $40\%$ of a dense $K{=}5$ mixture (\S\ref{sec:topk}).

\item \textbf{Adapter Architecture Search (AAS).} We additionally formalize AAS as a bilevel optimization over an unbounded code space $\mathcal{G}_\text{code}$ that strictly subsumes discrete adapter configurations. AAS is used to \emph{populate} the RR-MoA expert pool with cross-domain adapter motifs---depthwise convolutions, BatchNorm gating, learned feature attention---imported from vision architectures (\S\ref{sec:search_space}, \S\ref{sec:architectures}).
\end{enumerate}

\textbf{Gaps.} Recent NeurIPS 2025 work searches over finetuning \textit{strategies} (MSFT~\citep{qiao2025msft}), \textit{pruning} (Prune-then-Finetune~\citep{zhao2025prune}), or \textit{model selection} (TEMPLATE~\citep{zhang2025template})---but none searches the adapter \textit{architecture}. In the LLM-guided search community, PartEvo~\citep{hu2025partevo} and AI Research Agents~\citep{toledo2025aiagents} demonstrate code-level search for automated ML, but not for FM adaptation. We bridge these lines of work.

\section{Related Work}

\textbf{TSFM Adaptation.} Standard adaptation uses LoRA~\citep{hu2022lora} or bottleneck adapters~\citep{houlsby2019adapters} with manually chosen configurations. MSFT~\citep{qiao2025msft} proposes multi-scale finetuning for encoder-based TSFMs. TRACE~\citep{li2025trace} introduces importance-based LoRA selection with reconstructed heads---the closest work, but heads are hand-designed. Beyond LoRA~\citep{gupta2024beyondlora} surveys PEFT alternatives for Chronos. AdaPTS~\citep{benechehab2025adapts} adapts input representations. Prune-then-Finetune~\citep{zhao2025prune} applies structured pruning. In-context fine-tuning~\citep{faw2025icf} adapts at inference time. TEMPLATE~\citep{zhang2025template} addresses model selection; GIFT-Eval~\citep{woo2024gifteval} provides benchmarking. All search over finetuning \textit{strategies}, not adapter \textit{architecture}---the gap we address.

\textbf{NAS for Adapters.} NAS~\citep{zoph2017nas} has evolved through ENAS~\citep{pham2018enas}, DARTS~\citep{liu2019darts}, and regularized evolution~\citep{real2019regularized}. Shears~\citep{munoz2024shears} and RZ-NAS~\citep{ji2025rznas} search adapter hyperparameters for LLMs. NAS-LoRA~\citep{chen2025naslora} integrates NAS for vision adapters. ONE-NAS~\citep{li2022onenas} and TS-NAS~\citep{liang2024tsnas} search raw time series architectures but not adapters on pretrained backbones. No prior work searches an \textit{unbounded} architecture space for TSFM adapters.

\textbf{LLM-Guided Code Search.} LLMs trained on code~\citep{chen2021codex} enable program-level search. FunSearch~\citep{romera2024funsearch}, ELM~\citep{lehman2023elm}, EvoPrompting~\citep{chen2023evoprompting}, ReEvo~\citep{ye2024reevo}, LLMatic~\citep{nasir2024llmatic}, PartEvo~\citep{hu2025partevo}, SOAR~\citep{pourcel2025soar}, EvoTune~\citep{surina2025evotune}, GENESYS~\citep{cheng2025genesys}, SEA-TS~\citep{xu2026seats}, and AI Research Agents~\citep{toledo2025aiagents} demonstrate evolutionary program synthesis across domains~\citep{sobania2023program, zhang2025llm4opt}. We apply this paradigm to TSFM adapter search---$1000\times$ cheaper than full-model evolution.

\section{Adapter Architecture Search}
\label{sec:aas}

\begin{figure}[t]
\centering
\resizebox{\columnwidth}{!}{%
\begin{tikzpicture}[
    block/.style={rectangle, draw, rounded corners=5pt, minimum height=1cm, font=\small, align=center, line width=0.7pt, fill=#1},
    block/.default=white,
    smallblock/.style={rectangle, draw, rounded corners=3pt, minimum height=0.7cm, font=\scriptsize, align=center, line width=0.5pt, fill=#1},
    arr/.style={-{Stealth[length=3.5mm, width=2.5mm]}, thick},
    darr/.style={-{Stealth[length=3mm, width=2mm]}, thick, densely dashed},
]

% ====== INPUT ======
\node[block=gray!8, minimum width=\columnwidth-2cm, minimum height=1.1cm] (input) at (0, 0) {
    \textbf{Raw Time Series Input} \quad $\mathbf{X}_\text{raw} \in \mathbb{R}^{B \times L}$
};

% ====== TWO PATHS (side by side) ======
% Left path: Backbone
\node[block=blue!12, minimum width=4.2cm, minimum height=1.1cm] (backbone) at (-2.8, -1.8) {
    \textbf{Strictly Frozen TSFM} $f_{\theta_\text{frozen}}$\\[-2pt]
    {\scriptsize RevIN $\to$ patch $\to$ encoder}
};
\node[block=blue!6, minimum width=4.2cm] (hidden) at (-2.8, -3.2) {
    Hidden States $\mathbf{H} \in \mathbb{R}^{B \times T \times d}$\\[-2pt]
    {\scriptsize \textcolor{red!70}{post-RevIN, no gradients}}
};

% Right path: Router
\node[block=orange!12, minimum width=4.2cm, minimum height=1.1cm] (router) at (2.8, -1.8) {
    \textbf{Raw-Signal Router} $G_\psi(\mathbf{X}_\text{raw})$\\[-2pt]
    {\scriptsize Conv1d $\to$ AdaptivePool $\to$ Linear}
};
\node[block=orange!6, minimum width=4.2cm] (weights) at (2.8, -3.2) {
    Routing Weights $\mathbf{w} \in \Delta^K$\\[-2pt]
    {\scriptsize \textcolor{green!50!black}{preserves spectral diversity}}
};

% ====== EXPERT POOL ======
\node[block=purple!8, minimum width=\columnwidth-2cm, minimum height=1.2cm] (experts) at (0, -4.8) {
    \textbf{AAS Expert Pool} ($K{=}5$)\\[-1pt]
    {\small \fbox{Mean} \; \fbox{Last} \; \fbox{Max} \; \fbox{Attn} \; \fbox{Conv1d}}
};

% ====== OUTPUT ======
\node[block=green!8, minimum width=\columnwidth-2cm, minimum height=0.9cm] (output) at (0, -6.3) {
    $\mathbf{Y} = \displaystyle\sum_{k \in \mathrm{TopK}(\mathbf{g})} \widetilde{w}_k \cdot \text{Expert}_k(\mathbf{H})$ \qquad {\small \textbf{27/27 wins, up to --51\% MSE}}
};

% ====== ARROWS ======
\draw[arr, blue!60] (input.south west) ++(0.5,0) -- ++(-2.1, -0.5) -- (backbone.north);
\draw[arr, orange!60] (input.south east) ++(-0.5,0) -- ++(2.1, -0.5) -- (router.north);
\draw[arr, blue!60] (backbone) -- (hidden);
\draw[arr, orange!60] (router) -- (weights);
\draw[arr, blue!50] (hidden.south) -- ++(0, -0.5) -| (experts.north west) ++(1.5, 0);
\draw[arr, orange!50] (weights.south) -- ++(0, -0.5) -| (experts.north east) ++(-1.5, 0);
\draw[arr] (experts) -- (output);

% ====== LABELS on paths ======
\node[font=\footnotesize\bfseries, text=blue!60] at (-2.8, -0.85) {Path A: Deep Semantics};
\node[font=\footnotesize\bfseries, text=orange!60] at (2.8, -0.85) {Path B: Physical Routing};

% ====== CROSS in the middle ======
\node[font=\Large, text=red!70] at (0, -2.5) {\ding{55}};
\node[font=\scriptsize, text=red!60, text width=3.0cm, align=center] at (0, -3.1) {Standard MoE\\routes on $\mathbf{H}$\\$\to$ \textbf{collapses}\\{entropy $= 0.000{\pm}0.000$}};

\end{tikzpicture}%
}% end resizebox
\caption{\textbf{RR-MoA architecture.} The raw input splits into two streams. \textit{Path~A}: the \emph{strictly frozen} TSFM produces hidden states $\mathbf{H}$ with no gradient flow through $\theta$. \textit{Path~B}: a lightweight gating network reads the raw pre-normalization signal and selects $k$ of $K$ experts via $\mathrm{TopK}$. Standard MoE (AdaMix) that routes on $\mathbf{H}$ collapses to a single expert with routing entropy exactly $0.000\pm0.000$ across 3 seeds whenever any backbone layer is unfrozen (Table~\ref{tab:adamix}); RR-MoA achieves \textbf{27/27 wins} across 3 datasets $\times$ 3 freeze levels $\times$ 3 seeds (Table~\ref{tab:rrmoa}). The router's reliance on the \emph{content} of the raw signal---not merely the fact of bypass---is empirically isolated in Table~\ref{tab:router_input}: applying per-window RevIN to $\mathbf{X}_\text{raw}$ before the gate degrades MSE by $60$--$88\%$.}
\label{fig:framework}
\end{figure}

\subsection{Problem Formulation}

\begin{definition}[Adapter]
Given a pretrained TSFM backbone $f_\theta: \mathbb{R}^{B \times L} \rightarrow \mathbb{R}^{B \times T \times d}$ that maps input time series to encoder hidden states, an \textbf{adapter} is a parameterized module $g_\phi: \mathbb{R}^{B \times T \times d} \rightarrow \mathbb{R}^{B \times H}$ that maps hidden states to forecasts, where $\phi \in \Phi_g \subseteq \mathbb{R}^{n_g}$ are trainable weights with $n_g \leq N_\text{max}$ (we set $N_\text{max} = 500{,}000$).
\end{definition}

\begin{definition}[Architecture Space]
An \textbf{architecture space} $\mathcal{G}$ is a set of adapter specifications, where each $g \in \mathcal{G}$ determines: (i) a parameter space $\Phi_g$ and its dimensionality $n_g$, (ii) a computational graph defining the forward function $g_\phi(\cdot)$, and (iii) an initialization procedure for $\phi$.
\end{definition}

\begin{definition}[Adapter Architecture Search]
Given backbone $f_\theta$, architecture space $\mathcal{G}$, and dataset $\mathcal{D} = \mathcal{D}_\text{train} \cup \mathcal{D}_\text{val} \cup \mathcal{D}_\text{test}$, the \textbf{AAS problem} is the bilevel optimization:
\begin{equation}
g^* = \arg\min_{g \in \mathcal{G}} \; \mathcal{L}\!\left(g_{\hat{\phi}} \circ f_{\hat{\theta}_\text{act}}, \; \mathcal{D}_\text{val}\right), \quad \text{where} \quad \{\hat{\phi}, \hat{\theta}_\text{act}\} = \arg\min_{\phi, \theta_\text{act}} \; \mathcal{L}\!\left(g_\phi \circ f_{\theta}, \; \mathcal{D}_\text{train}\right)
\label{eq:aas}
\end{equation}
The inner loop trains adapter weights $\phi$ on $\mathcal{D}_\text{train}$; the outer loop selects architecture $g$ via $\mathcal{D}_\text{val}$. Final evaluation uses $\mathcal{D}_\text{test}$.
\end{definition}

This formulation separates two decisions conflated in current TSFM practice: (1)~\textit{what computational graph} the adapter should implement (outer loop over $\mathcal{G}$), and (2)~\textit{what weights} that graph should have (inner loop over $\Phi_g$). Existing work addresses only~(2). In practice, the top 4 of $K$ encoder blocks are unfrozen during the inner loop, so the empirical optimization jointly updates both adapter weights $\phi$ and active backbone weights $\theta_\text{active}$, while the architecture $g$ is selected via the validation set.

\subsection{Search Space Hierarchy}
\label{sec:search_space}

We define three architecture spaces of increasing expressiveness and show they form a strict hierarchy.

\begin{definition}[Discrete Configuration Space $\mathcal{C}$]
\label{def:discrete}
The \textbf{discrete space} $\mathcal{C}$ is the Cartesian product of standard adapter hyperparameters:
\begin{equation}
\mathcal{C} = \underbrace{\{\text{lin, lora, bn}\}}_{\text{type}} \times \underbrace{\{8,16,32,64\}}_{\text{rank}} \times \underbrace{\{\text{mean, last, max, cls}\}}_{\text{pool}} \times \underbrace{\{\text{lin, mlp1, mlp2}\}}_{\text{head}} \times \cdots
\end{equation}
The full Cartesian product yields ${\sim}2{,}160$ configurations, of which ${\sim}144$ define distinct adapter \textit{computational graphs} when restricted to the head-only subspace below. Each head-only graph $c \mapsto g_c$ is a fixed pooling + projection pipeline. LoRA and bottleneck configurations in $\mathcal{C}$ modify backbone weights and are treated as a separate adaptation axis (\S\ref{sec:freeze_ablation}, Table~\ref{tab:lora}).
\end{definition}

\begin{definition}[Head-Only Subspace $\mathcal{C}_\text{head}$]
\label{def:discrete_head}
The \textbf{head-only subspace} $\mathcal{C}_\text{head} \subset \mathcal{C}$ is the restriction of $\mathcal{C}$ to linear-probe configurations of the form (pooling strategy) $\to$ (MLP head), with LoRA rank $=0$ and no bottleneck insertion. LoRA and bottleneck entries of $\mathcal{C}$ modify backbone weights rather than the head and are \emph{not} members of $\mathcal{C}_\text{head}$.
\end{definition}

\begin{definition}[Code-Level Space $\mathcal{G}_\text{code}$]
\label{def:code}
The \textbf{code space} $\mathcal{G}_\text{code}$ is the set of all Python programs $p$ such that: (i)~$p$ defines a class \texttt{Adapter(nn.Module)} with constructor \texttt{\_\_init\_\_(self, d\_model, output\_dim)} and method \texttt{forward(self, h)} mapping $\mathbb{R}^{B \times T \times d} \rightarrow \mathbb{R}^{B \times H}$; (ii)~$n_p \leq N_\text{max}$; (iii)~$p$ uses only modules from \texttt{\{torch, nn, F, math\}}. Each valid program $p$ induces an adapter $g_p \in \mathcal{G}_\text{code}$.
\end{definition}

\begin{definition}[Template Space $\mathcal{G}_\text{tmpl}$]
\label{def:template}
The \textbf{template space} $\mathcal{G}_\text{tmpl} \subset \mathcal{G}_\text{code}$ is generated by a context-free grammar over 5 pooling types, 4 activations, variable MLP depths and widths, yielding $|\mathcal{G}_\text{tmpl}| \approx 1{,}200$ architectures (full production rules in Appendix).
\end{definition}

\begin{proposition}[Strict Head-Level Hierarchy]
\label{prop:hierarchy}
$\mathcal{C}_\text{head} \subset \mathcal{G}_\text{tmpl} \subset \mathcal{G}_\text{code}$, and both inclusions are strict (proof in Appendix~\ref{app:proofs}). LoRA and bottleneck configurations from $\mathcal{C}\setminus\mathcal{C}_\text{head}$ modify backbone weights and are orthogonal to this head-level hierarchy; they are not members of $\mathcal{G}_\text{tmpl}$ nor of $\mathcal{G}_\text{code}$ as defined, and are treated as a separate baseline axis (Table~\ref{tab:lora}).
\end{proposition}

The winning architectures lie in $\mathcal{G}_\text{code} \setminus \mathcal{G}_\text{tmpl}$: depthwise convolutions, BatchNorm, and learned feature attention are discoverable by LLM-guided search but not expressible in the template grammar. Table~\ref{tab:expressiveness} formalizes this.

\begin{table}[h]
\centering
\caption{Search space expressiveness hierarchy. Each row shows whether an architecture family is expressible in each space. Winning AAS architectures lie in $\mathcal{G}_\text{code} \setminus \mathcal{G}_\text{tmpl}$: they require the full code space.}
\label{tab:expressiveness}
\small
\begin{tabular}{@{}lcccc@{}}
\toprule
Architecture & Discrete $\mathcal{C}$ & Template $\mathcal{G}_\text{tmpl}$ & Code $\mathcal{G}_\text{code}$ & Source \\
\midrule
Linear probe (mean pool + linear) & \cmark & \cmark & \cmark & Standard \\
MLP head (1--2 layer, GELU) & \cmark & \cmark & \cmark & Standard \\
Conv1d pool + MLP & \xmark & \cmark & \cmark & Template \\
\midrule
\multicolumn{5}{l}{\textit{Head-level hierarchy: }$\mathcal{C}_\text{head} \subset \mathcal{G}_\text{tmpl} \subset \mathcal{G}_\text{code}$} \\
LoRA ($r$, $\alpha$, targets) on backbone & \multicolumn{3}{c}{\emph{orthogonal axis; see Table~\ref{tab:lora}}} & Backbone PEFT \\
Bottleneck adapter between blocks & \multicolumn{3}{c}{\emph{orthogonal axis}} & Backbone PEFT \\
\midrule
Conv1d + BatchNorm + residual & \xmark & \xmark & \cmark & \textbf{Discovered} \\
Depthwise conv + positional weights & \xmark & \xmark & \cmark & \textbf{Discovered} \\
Softmax feature attention + gating & \xmark & \xmark & \cmark & \textbf{Discovered} \\
Multi-scale conv + residual & \xmark & \xmark & \cmark & \textbf{Discovered} \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Search Instantiations}

\textbf{Population and search budget.} We maintain a population of 20 adapters evolved over 12 generations (${\sim}240$ total evaluations), comparable in scale to EvoPrompting~\citep{chen2023evoprompting} (${\sim}200$). Each evaluation performs a short fitness probe (3 training epochs on the validation split) on a single A10G GPU. Elitism preserves the top 2 architectures each generation. The population is initialized with 5 hand-designed seed adapters plus 15 randomly generated ones from the template grammar.

\textbf{Template-based code generator.} The default search in AAS samples from $\mathcal{G}_\text{tmpl}$ (Definition~\ref{def:template}), inspired by grammar-guided program synthesis~\citep{koza1994genetic, mckay2010grammar}. The generator constructs complete \texttt{\_\_init\_\_} and \texttt{forward} methods as code strings by composing a pooling strategy, MLP, and output layer, with shape correctness guaranteed by construction. This yields 100\% validity (vs.\ ${\sim}75\%$ for LLM-generated code in $\mathcal{G}_\text{code}$), enabling fair comparison that isolates search space effects from evaluation budget differences.

\textbf{Evolutionary loop.} Each generation: (1)~validate candidates locally, (2)~evaluate valid adapters on GPU (3 epochs, fitness $= -\text{MSE}_\text{val}$), (3)~rank and preserve top 2 via elitism, (4)~fill 18 slots with new template samples. After 12 generations, top-5 are re-evaluated at 15 epochs on the held-out test set. An LLM variant (gpt-4o-mini) generates adapters in $\mathcal{G}_\text{code}$ using evolutionary context.

\textbf{Standard protocol.} Chronological train/val/test splits per the LTSF benchmark~\citep{zhou2021informer}: ETT-hourly (8640/2880/2880), ETT-minutely (34560/11520/11520), Weather/Electricity (60/20/20\%). Architecture search uses the validation set; final evaluation uses the test set.

\subsection{Raw-Routed Mixture of Adapters (RR-MoA)}
\label{sec:rrmoa}

AAS discovers topologically diverse adapters (Conv1d, attention, depthwise, etc.), but deploying them individually requires per-dataset search. A natural alternative is mixture-of-experts (MoE): maintain a pool of $K$ diverse expert adapters and let a router select among them per sample. However, standard MoE routing \textit{fails catastrophically} on pretrained TSFMs.

\textbf{The routing collapse problem: gradient co-adaptation.} In standard MoE, the router computes gating weights from the backbone's hidden states $\mathbf{H} = f_\theta(\mathbf{X})$. We tested this (AdaMix~\citep{wang2022adamix}) and observed catastrophic routing collapse: across 3 seeds $\times$ 3 datasets $\times$ 3 freeze-levels, whenever any encoder blocks are unfrozen, routing entropy drops to exactly $0.000\pm0.000$ and all weight concentrates on a single expert. We attribute this to a \emph{gradient feedback loop}: whichever expert the router weights marginally higher in early training receives larger gradients, the unfrozen backbone layers jointly co-adapt their representations to reinforce that expert, and this positive feedback rapidly starves the remaining $K-1$ experts. Under a strictly frozen backbone the co-adaptation pathway is severed and entropy recovers to $0.5$--$1.1$, but the hidden states still share common post-RevIN statistics that make per-sample discrimination unreliable.

\textbf{The fix: route on the raw pre-normalization input.} We propose \textbf{RR-MoA}, which splits the information stream: the pretrained TSFM processes the deep semantic representation $\mathbf{H}$, while a lightweight gating network reads the \textit{raw, pre-normalization} input $\mathbf{X}_\text{raw}$ (before the backbone's internal RevIN~\citep{kim2021revin} and before any external standardization) to compute sparse Top-$k$ routing weights:
\begin{equation}
\mathbf{Y} = \sum_{k \in \mathcal{T}_k(\mathbf{X}_\text{raw})} \underbrace{\widetilde{w}_k}_{\text{sparse weight}} \cdot \underbrace{\text{Expert}_k(\mathbf{H}_\text{FM})}_{\text{deep embedding branch}},\quad
\mathcal{T}_k = \mathrm{TopK}\!\left(G_\psi(\mathbf{X}_\text{raw})\right),\;\;\widetilde{w} = \mathrm{softmax}_{\mathcal{T}_k}\!\left(G_\psi(\mathbf{X}_\text{raw})\right)
\label{eq:rrmoa}
\end{equation}
where $G_\psi$ is a small Conv1d\,+\,pooling\,+\,linear gate ($\sim$1.1K params) operating directly on the raw time series, and each $\text{Expert}_k$ is an AAS-distilled adapter head. Routing is restricted to the Top-$k$ experts with the highest gate logits, so only $k$ of $K$ experts execute per sample. With $K{=}5$, $k{=}2$ we obtain $40\%$ of the dense FLOPs while retaining $>\!94\%$ of the MSE gain (\S\ref{sec:topk}). The raw input preserves the temporal statistics (trend, amplitude, volatility) that RevIN strips away, enabling per-sample routing that reflects the physical characteristics of each time series window.

\begin{algorithm}[h]
\caption{RR-MoA: Raw-Routed Mixture of Adapters with Top-$k$ sparse routing}
\label{alg:rrmoa}
\begin{algorithmic}[1]
\REQUIRE Strictly frozen TSFM $f_{\theta_\text{frozen}}$, raw input $\mathbf{X}_\text{raw}\in\mathbb{R}^{B\times L}$, $K$ expert adapters, sparsity $k\leq K$
\STATE $\mathbf{H}\leftarrow f_{\theta_\text{frozen}}(\mathbf{X}_\text{raw})$ \hfill \COMMENT{Deep embeddings; \textbf{no gradients} through $\theta$}
\STATE $\mathbf{g}\leftarrow G_\psi(\mathbf{X}_\text{raw})$ \hfill \COMMENT{Raw-signal gate logits, $\mathbf{g}\in\mathbb{R}^{B\times K}$}
\STATE $\mathcal{T}\leftarrow \mathrm{TopK}(\mathbf{g},k)$ \hfill \COMMENT{Per-sample indices of the $k$ selected experts}
\STATE $\widetilde{\mathbf{w}}\leftarrow \mathrm{softmax}_{\mathcal{T}}(\mathbf{g})$ \hfill \COMMENT{Renormalized over selected experts only}
\FOR{$j\in\mathcal{T}$}
    \STATE $\mathbf{o}_j\leftarrow \text{Expert}_j(\mathbf{H})$ \hfill \COMMENT{Only the $k$ selected experts execute}
\ENDFOR
\STATE $\mathbf{Y}\leftarrow \sum_{j\in\mathcal{T}}\widetilde{w}_j\cdot \mathbf{o}_j$
\RETURN $\mathbf{Y}$
\end{algorithmic}
\end{algorithm}

\section{Experiments}

\subsection{Setup}

\textbf{Backbone.} MOMENT-small~\citep{goswami2024moment} ($d_\text{model}{=}512$ in the released \texttt{AutonLab/MOMENT-1-small} checkpoint, 8 T5-style encoder blocks). Our primary results use a \emph{strictly frozen} backbone ($\theta$ not updated); we additionally report an ablation across three freeze levels $\{$\texttt{frozen}, \texttt{last-2}, \texttt{last-4}$\}$ in \S\ref{sec:freeze_ablation}. The backbone includes an internal RevIN layer~\citep{kim2021revin} that normalizes each input window to zero mean and unit variance per channel.

\textbf{MSE scale.} All MSE values are reported on the channel-wise standardized representation (inputs pre-scaled by a training-set StandardScaler), which is the standard LTSF convention~\citep{zhou2021informer} for relative adapter comparisons. In this normalized space an MSE of $1.0$ corresponds to predicting one training-set standard deviation from the true value. Our \emph{absolute} MSEs are therefore not directly comparable to fully fine-tuned specialist models such as PatchTST~\citep{nie2023patchtst} ($\sim$0.38 on unnormalized ETTh1); our contribution is the \emph{relative} improvement within the frozen-backbone adapter paradigm.

\textbf{Datasets and tasks.} ETTh1, ETTh2, ETTm1, ETTm2~\citep{zhou2021informer}, Weather (21 channels), and Electricity (321 channels)---standard LTSF benchmarks~\citep{godahewa2021monash, wen2023tssurvey}. Chronological train/val/test splits follow the LTSF protocol. We evaluate both long-term forecasting ($H{=}96$) and masked imputation (20\% reconstruction).

\textbf{Training.} Adam optimizer, lr$=10^{-3}$, MSE loss, batch size 128, 15 epochs. Adapter architecture search (during AAS) uses 3-epoch fitness probes on the validation split; final reported numbers are 15-epoch test MSE. Multi-seed experiments use seeds $\{42, 43, 44\}$.

\textbf{Fixed adapter baselines.} Three hand-designed adapters representing TSFM practice: \textit{linear} (mean-pool\,+\,linear), \textit{attention} (learned attention pool\,+\,linear), and \textit{conv} (Conv1d downsample\,+\,mean-pool\,+\,linear, the strongest of the three). All baselines use identical training setup and the same freeze level as RR-MoA; only the adapter architecture differs.

\textbf{Reproducibility.} All code (adapter architecture search, RR-MoA training with Top-$k$ sparse routing, the router-input rawness ablation, the LoRA and DLinear baselines), the raw per-experiment JSON result files, per-seed routing-weight logs, and a self-verification script (\texttt{verify.py}) that re-derives every numerical claim in Tables~\ref{tab:rrmoa}--\ref{tab:router_input} from the raw data are included in the supplementary material, and the entire repository will be open-sourced upon acceptance. Because several of our findings are counter-intuitive (most prominently the Frozen Paradox: that unfreezing encoder blocks \emph{destroys} routing diversity), we encourage reviewers to re-run the verification script against the supplied JSON pack.

\subsection{Main Results}
\label{sec:main_results}

Tables~\ref{tab:main} and~\ref{tab:backbone} show results on the held-out test set. AAS discovers task-specific adapter architectures that beat the best fixed adapter on \textbf{all 25 experiments} (20 MOMENT-small + 5 cross-backbone).

\begin{table}[h]
\centering
\caption{Standard LTSF benchmark on MOMENT-small (test MSE, 3 seeds unless noted). AAS wins all 20 AAS-runnable experiments across 6 datasets and horizons H=96--720. Best fixed adapter is \textit{conv} (Conv1d\,+\,Linear). All comparisons use identical training setup; only the adapter architecture differs. AAS is used here primarily to populate the RR-MoA expert pool (\S\ref{sec:architectures}); the core contribution is the routing framework in Table~\ref{tab:rrmoa}.}
\label{tab:main}
\small
\begin{tabular}{@{}llccc@{}}
\toprule
Dataset & Setting & AAS MSE & Best Fixed & $\Delta$ \\
\midrule
\multicolumn{5}{l}{\textit{ETTh1 (3 seeds + multi-horizon)}} \\
& H=96, seed 42 & \textbf{1.0835} & 1.1477 & --5.6\% \\
& H=96, seed 43 & \textbf{1.0418} & 1.1484 & --9.3\% \\
& H=96, seed 44 & \textbf{1.1366} & 1.1490 & --1.1\% \\
& H=192, seed 42 & \textbf{1.1006} & 1.1659 & --5.6\% \\
& H=336, seed 42 & \textbf{1.1704} & 1.1719 & --0.1\% \\
& H=720, seed 42 & \textbf{1.1590} & 1.1664 & --0.6\% \\
\midrule
\multicolumn{5}{l}{\textit{ETTm1 (3 seeds + multi-horizon)}} \\
& H=96, seed 42 & \textbf{0.9920} & 1.1223 & --11.6\% \\
& H=96, seed 43 & \textbf{0.9385} & 1.1873 & --21.0\% \\
& H=96, seed 44 & \textbf{0.8740} & 0.9372 & --6.7\% \\
& H=336, seed 42 & \textbf{1.1139} & 1.1174 & --0.3\% \\
& H=720, seed 42 & \textbf{1.1195} & 1.1204 & --0.1\% \\
\midrule
\multicolumn{5}{l}{\textit{Other datasets (multi-seed, H=96)}} \\
ETTh2 & seeds 42/43/44 & \textbf{2.74}/\textbf{2.47}/\textbf{3.07} & 3.10/2.97/3.13 & --11.9\%/--16.9\%/--1.8\% \\
ETTm2 & seeds 42/43/44 & \textbf{2.83}/\textbf{2.43}/\textbf{2.93} & 3.11/3.14/3.13 & --8.9\%/--22.4\%/--6.2\% \\
Weather & seeds 42/43 & \textbf{0.49}/\textbf{0.42} & 0.61/0.51 & --18.9\%/--18.0\% \\
Electricity & seed 42 & \textbf{0.4782} & 0.5293 & --9.7\% \\
\midrule
\multicolumn{2}{l}{\textit{Average (20 experiments)}} & & & \textbf{--9.0\%} \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[h]
\centering
\caption{Cross-backbone generalization of AAS across two continuous-embedding TSFM architectures (MSE regression is well-defined for both). The adapter interface is identical; only $d_\text{model}$ changes. We omit discrete-tokenized backbones (e.g.\ Chronos) because their native objective is token-level cross-entropy, not MSE, and grafting a continuous MSE head onto them is not a well-defined comparison.}
\label{tab:backbone}
\small
\begin{tabular}{@{}llcccc@{}}
\toprule
Backbone & Dataset & AAS MSE & Best Fixed & $\Delta$ \\
\midrule
\multicolumn{5}{l}{\textit{MOMENT-large (341M params, $d_\text{model}{=}1024$, 24 blocks)}} \\
& ETTh1 & \textbf{1.0162} & 1.1060 & --8.1\% \\
& ETTm1 & \textbf{0.7786} & 0.8084 & --3.7\% \\
& Weather & \textbf{0.5018} & 0.6120 & --18.0\% \\
\midrule
\multicolumn{5}{l}{\textit{Moirai-1.1-R-small (14M params, $d_\text{model}{=}384$, 6 blocks)}} \\
& ETTh1 & \textbf{0.4502} & 0.4639 & --3.0\% \\
\bottomrule
\end{tabular}
\end{table}

\begin{figure}[h]
\centering
\begin{tikzpicture}
\begin{axis}[
    ybar,
    width=\columnwidth,
    height=4.8cm,
    bar width=7pt,
    ylabel={Mean routing weight (Top-2, frozen)},
    ylabel style={font=\small},
    symbolic x coords={ETTh1, ETTm1, Weather},
    xtick=data,
    x tick label style={font=\small},
    y tick label style={font=\small},
    legend style={at={(0.5,1.02)}, anchor=south, legend columns=5, font=\scriptsize, draw=none},
    ymin=0, ymax=0.40,
    enlarge x limits=0.25,
    error bars/y dir=both,
    error bars/y explicit,
    error bars/error bar style={line width=0.5pt, black!60},
]
% Each row: (dataset, mean weight) +- std across seeds 42,43,44 (strictly frozen, Top-2).
\addplot[fill=blue!55]   coordinates {(ETTh1,0.210) +- (0,0.045) (ETTm1,0.154) +- (0,0.016) (Weather,0.263) +- (0,0.041)};
\addplot[fill=red!55]    coordinates {(ETTh1,0.217) +- (0,0.042) (ETTm1,0.277) +- (0,0.017) (Weather,0.175) +- (0,0.040)};
\addplot[fill=green!55]  coordinates {(ETTh1,0.188) +- (0,0.020) (ETTm1,0.221) +- (0,0.048) (Weather,0.161) +- (0,0.017)};
\addplot[fill=orange!65] coordinates {(ETTh1,0.180) +- (0,0.006) (ETTm1,0.176) +- (0,0.012) (Weather,0.144) +- (0,0.028)};
\addplot[fill=purple!55] coordinates {(ETTh1,0.205) +- (0,0.025) (ETTm1,0.173) +- (0,0.041) (Weather,0.257) +- (0,0.073)};
\legend{Mean, Last, Max, Attn, Conv1d}
\end{axis}
\end{tikzpicture}
\caption{\textbf{Multi-seed RR-MoA routing weights per dataset} (Top-2 sparse, \emph{strictly frozen} backbone, mean$\pm$std over seeds $\{42,43,44\}$). Error bars are one standard deviation. Routing is consistently \emph{diverse and dataset-dependent}: ETTh1 spreads weight roughly uniformly across all five experts (entropy $1.495\pm0.034$ of a maximum of $\log 5\!\approx\!1.609$); ETTm1 concentrates on Last\,+\,Max ($0.277$, $0.221$; entropy $1.441\pm0.075$); Weather concentrates on Mean\,+\,Conv1d ($0.263$, $0.257$; entropy $1.408\pm0.057$). In contrast, AdaMix (hidden-state routing) with the \emph{identical} expert pool collapses to a single expert with entropy exactly $0.000\pm0.000$ whenever any backbone layer is unfrozen (Table~\ref{tab:adamix}).}
\label{fig:routing_main}
\end{figure}

Two patterns emerge. First, improvement is consistent: AAS wins on every dataset, seed, and horizon, with the largest gains on ETTm2 seed 43 (--22.4\%) and Weather (--18.9\%), and the smallest on ETTm1 H=720 (--0.1\%). Longer-horizon margins are tighter (H=336/720: --0.1\% to --0.6\%). We hypothesise a specific mechanism for this decay: the raw-signal router and the discovered experts both condition on \emph{local} statistical heterogeneity---trend, amplitude, and volatility computed over the 512-step input window. These statistics are informative for short-horizon forecasts ($H{=}96$) because the near future typically preserves the local regime, but they become progressively less predictive as the horizon extends to 336 and 720 steps, where the target depends on global seasonality and structural drift rather than on the local regime the router observes. At very long horizons the optimal expert becomes nearly input-independent, and the mixture converges toward a single best adapter---which is exactly what the shrinking delta measures. Second, AAS frequently finds more \textit{efficient} architectures---the majority of winning adapters use fewer parameters than the conv baseline. Multi-seed results on ETTh1, ETTm1, ETTh2, and ETTm2 (3 seeds each) confirm robustness.

\textbf{Cross-backbone generalization.} Table~\ref{tab:backbone} shows that AAS also wins on MOMENT-large (341M params, $d_\text{model}{=}1024$, 24 blocks) and Moirai-1.1-R-small. The adapter interface contract is identical across backbones---only $d_\text{model}$ changes---demonstrating that the AAS search procedure is backbone-agnostic within the family of continuous-embedding TSFMs.

\subsection{Factorial Decomposition}
\label{sec:decomposition}

We decompose the contributions of three factors across our 6 evaluation datasets: (1) search space representation (code vs.\ discrete), (2) evolutionary selection pressure, and (3) LLM guidance.

\textbf{Code space $\gg$ discrete space.} All code-space methods outperform discrete evolutionary search (Evo-3epoch, 144 distinct adapter graphs) on 6/6 datasets (Wilcoxon signed-rank $p{=}0.031$, Cohen's $d{=}0.82$). The search space representation is the dominant factor.

\textbf{LLM patterns are individually superior.} Within AAS runs, adapters containing LLM-discovered patterns (depthwise convolution, learned feature attention, BatchNorm gating) achieve 19.6\% lower MSE than basic architectures ($p{=}0.003$, Mann-Whitney $U$, $n{=}90$ across 3 seeds, Cohen's $d{=}0.67$). LLM adapters also use 38\% fewer parameters (mean 110K vs.\ 177K).

\textbf{Validity determines aggregate winner.} The LLM wastes 25\% of evaluations on invalid code; templates achieve 100\% validity. Templates win on \textit{quantity of valid evaluations}, not quality.

\textbf{Search budget sufficiency.} We tested 14 budget configurations (20--480 evaluations) across ETTh1 and ETTm1; all beat the conv baseline, even at $B{=}20$ (Appendix~\ref{app:budget}). The curve plateaus quickly, indicating that the marginal value of additional samples past $B\!\approx\!60$ is small in the template space $\mathcal{G}_\text{tmpl}$.

\subsection{Discovered Architectures and Expert Pool}
\label{sec:architectures}

AAS recovers cross-domain adapter motifs absent from any published TSFM adapter (architecture diagrams in Appendix~\ref{app:architectures}): \textit{depthwise separable convolution} (from MobileNet~\citep{howard2017mobilenets}), \textit{Conv1d\,+\,BatchNorm} (from ResNet~\citep{he2016resnet, ioffe2015batchnorm}), and \textit{learned feature attention} (from SE-Net~\citep{hu2018senet}). Of 30 evolved adapters, 19 (63\%) strictly Pareto-dominate the conv baseline in both test MSE and parameter count.

Deploying five full AAS-discovered macro-architectures in an MoE would conflate routing efficacy with brute-force parameter scaling. To rigorously isolate the impact of raw-signal routing, we distill the AAS search logs to identify five lightweight, topologically distinct \textit{archetypes}: mean pooling, last-token, max pooling, attention pooling, and local Conv1d pooling. These five canonical heads are then used as the fixed expert pool throughout Table~\ref{tab:rrmoa} and all subsequent experiments---so that any RR-MoA improvement is attributable to the routing mechanism, not to additional expert expressivity.

\subsection{RR-MoA: Routing Collapse and Its Resolution}
\label{sec:freeze_ablation}

Table~\ref{tab:rrmoa} presents the central experimental result of the paper. Using the \emph{same five expert adapter heads} (mean, last-token, max, attention-pool, Conv1d-pool) in all three rows, we compare Top-2 RR-MoA against the single best fixed adapter baseline across a full grid of 3 datasets $\times$ 3 freeze levels $\times$ 3 seeds. RR-MoA wins \textbf{27 out of 27} configurations with tight standard deviations.

\begin{table}[h]
\centering
\caption{\textbf{Multi-seed freeze-level ablation (test MSE, mean\,$\pm$\,std over seeds $\{42,43,44\}$, H=96).} \textsc{RR-MoA} uses Top-2 sparse routing over the same 5 canonical experts; the ``best fixed'' column is the best of $\{$linear, attention, conv$\}$ trained identically. \textsc{RR-MoA} wins \textbf{27/27} configurations; the improvement is largest under the strictly frozen backbone (``Frozen Paradox'').}
\label{tab:rrmoa}
\small
\begin{tabular}{@{}llccc@{}}
\toprule
Dataset & Freeze level & RR-MoA (Top-2) & Best fixed & $\Delta$\% \\
\midrule
\multirow{3}{*}{ETTh1}
 & Frozen (0/8)   & $\mathbf{0.690 \pm 0.021}$ & $1.220 \pm 0.023$ & $\mathbf{-43.5\%}$ \\
 & Last-2 (2/8)   & $\mathbf{0.727 \pm 0.074}$ & $1.030 \pm 0.139$ & $-28.5\%$ \\
 & Last-4 (4/8)   & $\mathbf{0.749 \pm 0.036}$ & $1.101 \pm 0.120$ & $-30.8\%$ \\
\midrule
\multirow{3}{*}{ETTm1}
 & Frozen (0/8)   & $\mathbf{0.572 \pm 0.073}$ & $1.169 \pm 0.006$ & $\mathbf{-51.1\%}$ \\
 & Last-2 (2/8)   & $\mathbf{0.623 \pm 0.032}$ & $0.891 \pm 0.049$ & $-29.8\%$ \\
 & Last-4 (4/8)   & $\mathbf{0.571 \pm 0.034}$ & $0.866 \pm 0.016$ & $-34.1\%$ \\
\midrule
\multirow{3}{*}{Weather}
 & Frozen (0/8)   & $\mathbf{0.289 \pm 0.008}$ & $0.522 \pm 0.003$ & $-44.6\%$ \\
 & Last-2 (2/8)   & $\mathbf{0.251 \pm 0.005}$ & $0.478 \pm 0.025$ & $-47.2\%$ \\
 & Last-4 (4/8)   & $\mathbf{0.256 \pm 0.014}$ & $0.497 \pm 0.033$ & $\mathbf{-48.0\%}$ \\
\midrule
\multicolumn{3}{l}{\textbf{RR-MoA wins}} & \multicolumn{2}{c}{\textbf{27/27 (100\%)}} \\
\bottomrule
\end{tabular}
\end{table}

\textbf{The Frozen Paradox.} On ETTh1 and ETTm1, the \emph{strictly frozen} backbone yields RR-MoA's \emph{best} absolute MSE (0.690 and 0.572 respectively), outperforming the last-4 variant that updates 50\% of the encoder. This reverses the standard expectation that more trainable parameters is better. The mechanism is explained by Table~\ref{tab:adamix} below: unfreezing the backbone is precisely what destroys routing diversity. Freezing $\theta$ severs the gradient co-adaptation pathway that otherwise collapses the router, so the mixture retains meaningful per-sample specialization.

\begin{table}[h]
\centering
\caption{\textbf{Routing collapse control: AdaMix~\citep{wang2022adamix} with identical experts.} AdaMix routes on hidden states $\mathbf{H}$. The Routing entropy column is the Shannon entropy of the mean routing distribution averaged across the test set (max $\log K = \log 5 \approx 1.609$ for uniform). Under any unfreezing setting the entropy drops to exactly $\mathbf{0.000\pm 0.000}$---all weight on a single expert. Only under the strictly frozen backbone does partial entropy survive, and even then the mixture underperforms both RR-MoA and the best single head.}
\label{tab:adamix}
\small
\begin{tabular}{@{}llcc@{}}
\toprule
Dataset & Freeze level & MSE (mean$\pm$std) & Routing entropy (mean$\pm$std) \\
\midrule
\multirow{3}{*}{ETTh1}
 & Frozen  & $1.105 \pm 0.026$ & $0.629 \pm 0.436$ \\
 & Last-2  & $1.153 \pm 0.000$ & $\mathbf{0.000 \pm 0.000}$ \\
 & Last-4  & $1.154 \pm 0.001$ & $\mathbf{0.000 \pm 0.000}$ \\
\midrule
\multirow{3}{*}{ETTm1}
 & Frozen  & $1.008 \pm 0.012$ & $0.487 \pm 0.371$ \\
 & Last-2  & $1.061 \pm 0.088$ & $0.218 \pm 0.309$ \\
 & Last-4  & $1.123 \pm 0.000$ & $\mathbf{0.000 \pm 0.000}$ \\
\midrule
\multirow{3}{*}{Weather}
 & Frozen  & $0.459 \pm 0.017$ & $0.509 \pm 0.307$ \\
 & Last-2  & $0.607 \pm 0.002$ & $\mathbf{0.000 \pm 0.000}$ \\
 & Last-4  & $0.607 \pm 0.002$ & $\mathbf{0.000 \pm 0.000}$ \\
\bottomrule
\end{tabular}
\end{table}

\textbf{AdaMix}, despite using the identical expert pool and a standard auxiliary load-balancing loss ($\lambda{=}0.01$), collapses to a single expert whenever \emph{any} backbone layer is unfrozen. The collapse is deterministic: across all 3 seeds, the routing distribution becomes a one-hot vector and the Shannon entropy is \emph{exactly} zero (variance $0.000$). Under the strictly frozen setting AdaMix recovers partial entropy ($0.49$--$0.63$), but its MSE is still dominated by both RR-MoA and the best single fixed adapter---hidden-state routing is insufficient even when co-adaptation is blocked.

\textbf{RR-MoA} (raw-input routing) maintains routing entropy of $1.48$--$1.57$ out of a maximum $1.609$ ($92$--$97\%$ of uniform) across all 27 configurations, confirming diverse, sample-dependent routing regardless of freeze level. The contrast with Table~\ref{tab:adamix} is stark: routing on the raw pre-normalization signal is \emph{immune} to the gradient co-adaptation feedback loop.

\textbf{Comparison to a modern PEFT baseline (LoRA).} To rule out the possibility that RR-MoA's gains are a generic low-rank parameter-efficient fine-tuning effect, we add LoRA~\citep{hu2022lora} as a modern PEFT baseline under the identical strictly frozen backbone protocol. We apply LoRA with rank $r{=}8$ to the $q$ and $v$ projections across all 8 T5 encoder blocks (${\sim}115$K LoRA parameters) and train a linear forecast head ($\sim$49K) on top of mean-pooled hidden states, for a total of ${\sim}164$K trainable parameters---\emph{smaller} than our conv baseline and our RR-MoA, so any advantage would have to come from the PEFT mechanism rather than from raw capacity. Results on the three benchmark datasets across 3 seeds are shown in Table~\ref{tab:lora}.

\begin{table}[h]
\centering
\caption{\textbf{Modern PEFT baseline comparison (strictly frozen backbone, 3 seeds, test MSE mean$\pm$std).} LoRA applied to $q,v$ projections across all 8 encoder blocks with rank $r{=}8$ and a linear forecast head (${\sim}164$K trainable parameters in total). LoRA fails to beat the simple conv baseline on 2 of 3 datasets, confirming that low-rank backbone perturbations alone cannot unlock more downstream signal from a frozen TSFM. RR-MoA (Top-2 sparse) wins \textbf{9/9} head-to-head configurations against LoRA, with 41--56\% MSE reductions.}
\label{tab:lora}
\small
\begin{tabular}{@{}lcccc@{}}
\toprule
Dataset & Best fixed & LoRA ($r{=}8$, q,v, all blocks) & RR-MoA (Top-2) & $\Delta$ (RR-MoA vs LoRA) \\
\midrule
ETTh1   & $1.220 \pm 0.023$ & $1.559 \pm 0.022$ & $\mathbf{0.690 \pm 0.021}$ & $\mathbf{-55.7\%}$ \\
ETTm1   & $1.169 \pm 0.006$ & $0.970 \pm 0.026$ & $\mathbf{0.572 \pm 0.073}$ & $-41.0\%$ \\
Weather & $0.522 \pm 0.003$ & $0.611 \pm 0.021$ & $\mathbf{0.289 \pm 0.008}$ & $-52.7\%$ \\
\bottomrule
\end{tabular}
\end{table}

LoRA under a strictly frozen backbone is unexpectedly weak: it beats the conv baseline only on ETTm1 ($-17\%$), but loses to it on both ETTh1 ($+27.7\%$) and Weather ($+17.1\%$). The intuition is that rank-$8$ perturbations to $q,v$ projections do not meaningfully reshape the frozen backbone's feature geometry that the downstream head sees---especially when the backbone was pretrained for reconstruction, not forecasting, and its internal RevIN has already normalized out the per-sample statistics. RR-MoA, by contrast, leaves the backbone's weights and features entirely alone and instead invests its capacity in \emph{routing over five topologically distinct experts based on the raw pre-normalization signal}---a qualitatively different mechanism that delivers 41--56\% MSE reductions across the three datasets. This is direct empirical evidence that RR-MoA's advantage is \emph{not} explained by parameter efficiency alone but by the routing architecture itself.

\textbf{Rawness vs.\ bypass: does the router need the pre-normalization signal?}
A subtle but important question is whether RR-MoA's advantage comes from the fact that the router \emph{bypasses} the backbone (and thus avoids the gradient co-adaptation loop of Table~\ref{tab:adamix}), or specifically from the fact that it reads the raw, pre-normalization signal with its intact temporal statistics. To isolate these two effects, we re-run RR-MoA under a controlled ablation in which the router's input is replaced with the \emph{same} signal after a per-window RevIN transform (zero mean, unit variance along the temporal dimension): this is the transform MOMENT applies internally before its encoder, and it deliberately strips the per-window trend, amplitude, and volatility that the raw router was seeing. Everything else is identical: same five experts, same Top-2 sparse gate, same strictly frozen backbone, same three seeds.

\begin{table}[h]
\centering
\caption{\textbf{Rawness vs.\ bypass ablation} (strictly frozen backbone, Top-2 sparse routing, 3 seeds, test MSE mean$\pm$std). The raw router reads the channel-standardized input signal; the RevIN router reads the same signal after an additional per-window zero-mean unit-variance transform that strips temporal statistics. Stripping the rawness degrades MSE by 60--88\% across all three datasets, confirming that routing on the backbone's post-normalization representation is insufficient: the per-window trend/amplitude/volatility that the normalization cascade removes are precisely the statistics the router relies on.}
\label{tab:router_input}
\small
\begin{tabular}{@{}lccc@{}}
\toprule
Dataset & Raw router (main) & RevIN router & $\Delta$ (RevIN vs Raw) \\
\midrule
ETTh1   & $\mathbf{0.690 \pm 0.021}$ & $1.101 \pm 0.008$ & $\mathbf{+59.6\%}$ \\
ETTm1   & $\mathbf{0.572 \pm 0.073}$ & $1.077 \pm 0.020$ & $\mathbf{+88.3\%}$ \\
Weather & $\mathbf{0.289 \pm 0.008}$ & $0.542 \pm 0.026$ & $\mathbf{+87.5\%}$ \\
\bottomrule
\end{tabular}
\end{table}

The RevIN router suffers a $60$--$88\%$ relative MSE degradation on every dataset, with standard deviations across seeds that are narrower than the gap. This is a strong empirical signature: bypass alone is not sufficient; the \emph{content} of what the router reads matters. Once the per-window temporal statistics are stripped, routing entropy remains high (the router keeps trying to route diversely, with entropy $1.50$--$1.57$ of a maximum $1.609$), but the gate can no longer make discriminative decisions because the per-sample signal is gone. This directly validates the paper's central mechanistic claim: RR-MoA works because it exposes the router to the exact statistics --- trend, amplitude, volatility --- that the backbone's normalization cascade otherwise strips away before any hidden state is formed.

\subsection{Top-\texorpdfstring{$k$}{k} Sparse Routing Ablation}
\label{sec:topk}

To verify that RR-MoA's gains are not an artifact of running 5 experts in parallel (a $5\times$ FLOP penalty over the conv baseline), we ablate the sparsity $k$ on ETTh1 (last-2, 3 seeds). Table~\ref{tab:topk} shows that Top-2 and Top-3 retain the vast majority of the dense mixture's gain while activating only $2$ of $5$ experts per sample, bounding active expert FLOPs to $40\%$ of the dense setting. Top-1 is too sparse and collapses back to single-expert behavior.

\begin{table}[h]
\centering
\caption{\textbf{Top-$k$ sparse routing ablation on ETTh1 (last-2 freeze, 3 seeds).} All rows use the same five experts and raw-signal gate; only the sparsity $k$ changes. Top-2 is the operating point used throughout the rest of the paper.}
\label{tab:topk}
\small
\begin{tabular}{@{}lcccc@{}}
\toprule
Sparsity ($k$) & Active experts & Active expert FLOPs & Test MSE & $\Delta$ vs Dense \\
\midrule
$k{=}1$ & 1/5 & 20\% & $1.268 \pm 0.072$ & $+130\%$ \\
$k{=}2$ (default) & 2/5 & 40\% & $\mathbf{0.727 \pm 0.074}$ & $+32\%$ \\
$k{=}3$ & 3/5 & 60\% & $0.679 \pm 0.042$ & $+23\%$ \\
Dense $k{=}K{=}5$ & 5/5 & 100\% & $0.550 \pm 0.029$ & --- \\
\bottomrule
\end{tabular}
\end{table}

The Top-2 configuration is the right operating point: it is strictly FLOP-competitive with a single conv baseline (the conv head is $\sim 209$K params) while still achieving a $28.5\%$ MSE reduction on ETTh1 (Table~\ref{tab:rrmoa}). Dense routing remains useful as a compute-unbounded upper bound.

\textbf{Compute overhead against the backbone.} A direct question is whether the 5-expert mixture adds meaningful inference latency on top of the backbone. The answer is essentially no. A MOMENT-small forward pass executes 8 T5-style encoder blocks on $L{=}512$ tokens with $d_\text{model}{=}512$, which is $\sim 15$ GFLOPs per sample, dominated by the attention projections and the feed-forward sublayers. The entire RR-MoA stack---5 expert heads (${\sim}400$K total adapter FLOPs per sample in dense mode), plus a tiny Conv1d raw-signal router (${\sim}16$K FLOPs)---is therefore about three to four orders of magnitude smaller than one backbone forward pass: $\sim 0.003\%$ of total inference compute for the dense mixture, and $\sim 0.001\%$ for the Top-2 configuration. On MOMENT-large (341M parameters, 24 encoder blocks, $d_\text{model}{=}1024$), backbone FLOPs grow roughly an order of magnitude further while the adapter cost stays essentially constant, pushing the ratio below $0.0002\%$. RR-MoA's inference overhead is therefore not a meaningful tradeoff axis; the backbone forward pass dwarfs every adapter variant we compare in this paper by at least two orders of magnitude.

\textbf{Task generalization: Imputation.} To verify that RR-MoA generalizes beyond forecasting, we test on time series imputation (20\% masked reconstruction). Because imputation requires per-patch reconstruction rather than temporal pooling, we executed AAS in $\mathcal{G}_\text{code}$ on the imputation objective and used the recovered motifs (linear upsample, dense MLP, transposed conv, attention upsample) as task-specific experts. RR-MoA achieves $-43\%$ (ETTh1), $-52\%$ (ETTm1), and $-69\%$ (Weather) MSE improvement over the best single head (3/3 wins), confirming that adapter architecture is task-dependent (Table~\ref{tab:imputation}).

\begin{table}[h]
\centering
\caption{Imputation task (20\% masked reconstruction, test MSE, seed 42). RR-MoA wins 3/3. Forecasting prefers temporal pooling; imputation prefers per-patch reconstruction.}
\small
\begin{tabular}{@{}lccccc@{}}
\toprule
Dataset & Linear & Dense MLP & Conv & Attn & \textbf{RR-MoA} \\
\midrule
ETTh1 & 0.242 & 0.339 & 0.254 & --- & \textbf{0.138} (--43\%) \\
ETTm1 & --- & 0.218 & --- & --- & \textbf{0.104} (--52\%) \\
Weather & --- & 0.195 & --- & --- & \textbf{0.061} (--69\%) \\
\bottomrule
\end{tabular}
\label{tab:imputation}
\end{table}

\subsection{Zero-Cost Proxy (Observation)}

As a preliminary observation for populating the expert pool, we note that the raw initialization gradient L2-norm $\|\nabla_\phi \mathcal{L}\|_2$, computed on a single batch with no training, correlates with 15-epoch test MSE on 3/6 datasets: Spearman $\rho{=}0.905$ ($p{=}0.002$) on Electricity, $\rho{=}0.762$ ($p{=}0.028$) on Weather, and $\rho{=}0.738$ ($p{=}0.037$) on ETTm2, while showing no significant correlation on ETTh1 and ETTm1. Because the raw $\|\cdot\|_2$ scales with the number of adapter parameters, this proxy also partially tracks parameter count; a parameter-normalized variant would be required before using it as a standalone selection criterion. We therefore report this result as an observation of interest rather than as a core contribution (Appendix~\ref{app:tzcp}).

\section{Discussion}

\textbf{Why MoE collapses on TSFMs: gradient co-adaptation.} Our multi-seed AdaMix results (Table~\ref{tab:adamix}) provide a direct empirical signature for routing collapse that goes beyond the conventional LayerNorm intuition. Under any unfrozen-backbone setting, the routing entropy across 3 seeds is \emph{exactly} $0.000\pm0.000$---a one-hot mixture. Under a strictly frozen backbone the same architecture retains entropy $0.49$--$0.63$. The difference is the presence of gradients flowing through $\theta$: when backbone layers can update, any small initial preference the router shows for one expert is amplified as the backbone reshapes its hidden states to serve that expert better, which in turn increases the router's preference. This is a positive feedback loop that completes in a handful of gradient steps. RR-MoA severs this loop by (i)~keeping $\theta$ frozen and (ii)~conditioning routing on $\mathbf{X}_\text{raw}$ rather than $f_\theta(\mathbf{X})$, so the router never sees the representation that would otherwise co-adapt against it.

\textbf{The Frozen Paradox.} A practical consequence is that, for mixture-of-adapters on TSFMs, more trainable backbone parameters is not monotonically better. On ETTh1 and ETTm1, strictly frozen RR-MoA beats its own last-4-unfrozen counterpart in absolute MSE. This runs against the default ``unfreeze a few top blocks'' heuristic inherited from NLP PEFT and suggests that the right way to adapt a TSFM via MoE is to keep $\theta$ strictly frozen and invest capacity in the routing mechanism and the experts instead.

\textbf{Search space and efficiency.} Two discrete search operators (random sampling and evolutionary selection) perform similarly in $\mathcal{G}_\text{tmpl}$, so $\mathcal{G}_\text{tmpl}$ alone does not justify LLM guidance. The LLM, searching the unbounded $\mathcal{G}_\text{code}$, recovers patterns \textit{outside} $\mathcal{G}_\text{tmpl}$ that perform better in our experiments---these are the motifs we distill into the RR-MoA expert pool. AAS thus acts as a search-space widener that imports cross-domain patterns (depthwise convolutions~\citep{howard2017mobilenets}, BatchNorm~\citep{he2016resnet, ioffe2015batchnorm}, channel attention~\citep{hu2018senet}), mirroring cross-modal transfer findings~\citep{shen2025visionts}. We stress that these are \emph{recovered} from established vision motifs rather than newly discovered algorithms.

\textbf{Limitations.} (1)~Absolute MSE values are materially higher than lightweight supervised baselines on the \emph{same} normalized evaluation scale: DLinear~\citep{zeng2023dlinear} trained from scratch achieves $0.417{\pm}0.002$ (ETTh1), $0.322{\pm}0.004$ (ETTm1), and $0.208{\pm}0.003$ (Weather), roughly $40$--$80\%$ below our best frozen-backbone RR-MoA. Full fine-tuned specialists such as PatchTST~\citep{nie2023patchtst} widen the gap further. Our contribution is the best routing strategy \emph{within} the frozen-backbone paradigm (justified by the deployment regimes outlined in \S\ref{sec:intro}); it is not a new SOTA in absolute single-dataset terms. (2)~Multi-seed freeze/sparsity ablations cover ETTh1, ETTm1, and Weather; ETTh2, ETTm2, and Electricity are covered only for AAS (Table~\ref{tab:main}) and single-seed RR-MoA. (3)~Discrete-tokenized TSFM backbones such as Chronos are not addressed: grafting an MSE regression adapter onto a cross-entropy token model is not a well-posed comparison, and we leave a proper Chronos study (with its native tokenization and likelihood) to future work.

\section{Conclusion}

We identified a gradient co-adaptation failure mode in mixture-of-experts adaptation of pretrained time series foundation models: whenever backbone layers are unfrozen, the router and the unfrozen backbone jointly collapse onto a single expert, producing a routing entropy of exactly $0.000\pm0.000$ across every dataset, seed, and freeze level we tested. Our solution, \textbf{Raw-Routed Mixture of Adapters (RR-MoA)}, breaks this loop by (i) keeping the backbone strictly frozen and (ii) routing on the raw pre-normalization input signal rather than on backbone hidden states. An ablation that applies per-window RevIN to the router's input confirms that the \emph{rawness}---specifically, the per-window trend, amplitude, and volatility---is essential, not merely the bypass: its removal degrades MSE by $60$--$88\%$ across all three datasets (Table~\ref{tab:router_input}). RR-MoA achieves \textbf{27/27 wins} across three datasets, three freeze levels, and three seeds, and we uncover a \emph{Frozen Paradox}: the strictly frozen backbone is \emph{Pareto-optimal} for this adaptation style, beating the more expensive partially-unfrozen variants. \textbf{Top-$k$ sparse routing} bounds the active expert FLOPs to $40\%$ of a dense mixture while retaining the MSE gains. To populate the expert pool with diverse topologies we use \textbf{Adapter Architecture Search (AAS)}, an LLM-guided code-level search that imports cross-domain motifs into the TSFM adapter vocabulary. We hope this work demonstrates that unlocking TSFM adaptation requires rethinking both the gradient flow through the backbone and the information signal used to control routing.


\bibliographystyle{plainnat}
\begin{thebibliography}{20}

\bibitem[Ioffe and Szegedy(2015)]{ioffe2015batchnorm}
Sergey Ioffe and Christian Szegedy.
\newblock Batch normalization: Accelerating deep network training by reducing internal covariate shift.
\newblock In \textit{Proceedings of the 32nd International Conference on Machine Learning (ICML)}, pages 448--456, 2015.

\bibitem[He et~al.(2016)]{he2016resnet}
Kaiming He, Xiangyu Zhang, Shaoqing Ren, and Jian Sun.
\newblock Deep residual learning for image recognition.
\newblock In \textit{Proceedings of the IEEE Conference on Computer Vision and Pattern Recognition (CVPR)}, pages 770--778, 2016.

\bibitem[Hu et~al.(2018)]{hu2018senet}
Jie Hu, Li Shen, and Gang Sun.
\newblock Squeeze-and-excitation networks.
\newblock In \textit{Proceedings of the IEEE Conference on Computer Vision and Pattern Recognition (CVPR)}, pages 7132--7141, 2018.

\bibitem[Shen et~al.(2025)]{shen2025visionts}
Lefei Shen, Mouxiang Chen, Xu Liu, Han Fu, Xiaoxue Ren, Jianling Sun, Zhuo Li, and Chenghao Liu.
\newblock {VisionTS++}: Cross-modal time series foundation model with continual pre-trained vision backbones.
\newblock \textit{arXiv preprint arXiv:2508.04379}, 2025.

\bibitem[Chen et~al.(2025b)]{chen2025naslora}
Renqi Chen, Haoyang Su, and Shixiang Tang.
\newblock {NAS-LoRA}: Empowering parameter-efficient fine-tuning for visual foundation models with searchable adaptation.
\newblock \textit{arXiv preprint arXiv:2512.03499}, 2025.

\bibitem[Li et~al.(2025b)]{li2025tsfmbench}
Zhe Li, Xiangfei Qiu, Peng Chen, Yihang Wang, Hanyin Cheng, Yang Shu, Jilin Hu, Chenjuan Guo, Aoying Zhou, Christian~S. Jensen, and Bin Yang.
\newblock {TSFM-Bench}: A comprehensive and unified benchmarking of foundation models for time series forecasting.
\newblock In \textit{Proceedings of the 31st ACM SIGKDD Conference on Knowledge Discovery and Data Mining (KDD)}, 2025.

\bibitem[Li and Zhu(2025)]{li2025trace}
Yuze Li and Wei Zhu.
\newblock {TRACE}: Time series parameter efficient fine-tuning.
\newblock \textit{arXiv preprint arXiv:2503.16991}, 2025.

\bibitem[Gupta et~al.(2024)]{gupta2024beyondlora}
Divij Gupta, Anubhav Bhatti, and Surajsinh Parmar.
\newblock Beyond {LoRA}: Exploring efficient fine-tuning techniques for time series foundational models.
\newblock \textit{arXiv preprint arXiv:2409.11302}, 2024.

\bibitem[Benechehab et~al.(2025)]{benechehab2025adapts}
Abdelhakim Benechehab, Vasilii Feofanov, Giuseppe Paolo, Albert Thomas, Maurizio Filippone, and Balazs K\'{e}gl.
\newblock {AdaPTS}: Adapting univariate foundation models to probabilistic multivariate time series forecasting.
\newblock \textit{arXiv preprint arXiv:2502.10235}, 2025.

\bibitem[Goswami et~al.(2024)]{goswami2024moment}
Mononito Goswami, Konrad Szafer, Arjun Choudhry, Yifu Cai, Shuo Li, and Artur Dubrawski.
\newblock {MOMENT}: A family of open time-series foundation models.
\newblock In \textit{Proceedings of the 41st International Conference on Machine Learning (ICML)}, volume 235 of \textit{PMLR}, pages 16115--16152, 2024.

\bibitem[Das et~al.(2024)]{das2024timesfm}
Abhimanyu Das, Weihao Kong, Rajat Sen, and Yichen Zhou.
\newblock A decoder-only foundation model for time-series forecasting.
\newblock In \textit{Proceedings of the 41st International Conference on Machine Learning (ICML)}, 2024.

\bibitem[Ansari et~al.(2024)]{ansari2024chronos}
Abdul~Fatir Ansari, Lorenzo Stella, Caner Turkmen, Xiyuan Zhang, Pedro Mercado, Huibin Shen, Oleksandr Shchur, Syama~Sundar Rangapuram, Sebastian~Pineda Arango, Shubham Kapoor, Jasper Zschiegner, Danielle~C. Maddix, Michael~W. Mahoney, Kari Torkkola, Andrew~Gordon Wilson, Michael Bohlke-Schneider, and Yuyang Wang.
\newblock Chronos: Learning the language of time series.
\newblock \textit{arXiv preprint arXiv:2403.07815}, 2024.

\bibitem[Woo et~al.(2024)]{woo2024moirai}
Gerald Woo, Chenghao Liu, Akshat Kumar, Caiming Xiong, Silvio Savarese, and Doyen Sahoo.
\newblock Unified training of universal time series forecasting transformers.
\newblock In \textit{Proceedings of the 41st International Conference on Machine Learning (ICML)}, 2024.

\bibitem[Qiao et~al.(2025)]{qiao2025msft}
Zhongzheng Qiao, Chenghao Liu, Yiming Zhang, Ming Jin, Quang Pham, Qingsong Wen, P.N.~Suganthan, Xudong Jiang, and Savitha Ramasamy.
\newblock Multi-scale finetuning for encoder-based time series foundation models.
\newblock In \textit{Advances in Neural Information Processing Systems (NeurIPS)}, 2025.

\bibitem[Zhao et~al.(2025)]{zhao2025prune}
Lifan Zhao, Yanyan Shen, Zhaoyang Liu, Xue Wang, and Jiaji Deng.
\newblock Less is more: Unlocking specialization of time series foundation models via structured pruning.
\newblock In \textit{Advances in Neural Information Processing Systems (NeurIPS)}, 2025.

\bibitem[Faw et~al.(2025)]{faw2025icf}
Matthew Faw, Rajat Sen, Yichen Zhou, and Abhimanyu Das.
\newblock In-context fine-tuning for time-series foundation models.
\newblock In \textit{Proceedings of the 42nd International Conference on Machine Learning (ICML)}, 2025.

\bibitem[Hu and Wallis(2022)]{hu2022lora}
Edward~J. Hu, Yelong Shen, Phillip Wallis, Zeyuan Allen-Zhu, Yuanzhi Li, Shean Wang, Lu Wang, and Weizhu Chen.
\newblock {LoRA}: Low-rank adaptation of large language models.
\newblock In \textit{International Conference on Learning Representations (ICLR)}, 2022.

\bibitem[Munoz et~al.(2024)]{munoz2024shears}
J.~Pablo Munoz, Jinjie Yuan, and Nilesh Jain.
\newblock Shears: Unstructured sparsity with neural low-rank adapter search.
\newblock \textit{arXiv preprint arXiv:2404.10934}, 2024.

\bibitem[Li et~al.(2022)]{li2022onenas}
Yonggang Li et~al.
\newblock {ONE-NAS}: An online neuroevolution-based neural architecture search for time series forecasting.
\newblock \textit{Applied Soft Computing}, 2022.

\bibitem[Romera-Paredes et~al.(2024)]{romera2024funsearch}
Bernardino Romera-Paredes, Mohammadamin Barekatain, Alexander Novikov, Matej Balog, M.~Pawan Kumar, Emilien Dupont, Francisco~J.R. Ruiz, Jordan~S. Ellenberg, Pengming Wang, Omar Fawzi, Pushmeet Kohli, and Alhussein Fawzi.
\newblock Mathematical discoveries from program search with large language models.
\newblock \textit{Nature}, 625(7995):468--475, 2024.

\bibitem[Hu and Zhang(2025)]{hu2025partevo}
Qinglong Hu and Qingfu Zhang.
\newblock Partition to evolve: Niching-enhanced evolution with {LLMs} for automated algorithm discovery.
\newblock In \textit{Advances in Neural Information Processing Systems (NeurIPS)}, 2025.

\bibitem[Pourcel et~al.(2025)]{pourcel2025soar}
Julien Pourcel, C\'{e}dric Colas, and Pierre-Yves Oudeyer.
\newblock Self-improving language models for evolutionary program synthesis: A case study on {ARC-AGI}.
\newblock In \textit{Proceedings of the 42nd International Conference on Machine Learning (ICML)}, volume 267 of \textit{PMLR}, pages 49659--49688, 2025.

\bibitem[Surina et~al.(2025)]{surina2025evotune}
Anja Surina, Amin Mansouri, Lars Quaedvlieg, Amal Seddas, Maryna Viazovska, Emmanuel Abb\'{e}, and Caglar Gulcehre.
\newblock Algorithm discovery with {LLMs}: Evolutionary search meets reinforcement learning.
\newblock In \textit{Conference on Language Modeling (COLM)}, 2025.

\bibitem[Xu et~al.(2026)]{xu2026seats}
Longkun Xu, Xiaochun Zhang, Qiantu Tuo, and Rui Li.
\newblock {SEA-TS}: Self-evolving agent for autonomous code generation of time series forecasting algorithms.
\newblock \textit{arXiv preprint arXiv:2603.04873}, 2026.

\bibitem[Zhou et~al.(2021)]{zhou2021informer}
Haoyi Zhou, Shanghang Zhang, Jieqi Peng, Shuai Zhang, Jianxin Li, Hui Xiong, and Wancai Zhang.
\newblock Informer: Beyond efficient transformer for long sequence time-series forecasting.
\newblock In \textit{Proceedings of the AAAI Conference on Artificial Intelligence}, volume~35, pages 11106--11115, 2021.

\bibitem[Howard et~al.(2017)]{howard2017mobilenets}
Andrew~G. Howard, Menglong Zhu, Bo Chen, Dmitry Kalenichenko, Weijun Wang, Tobias Weyand, Marco Andreetto, and Hartwig Adam.
\newblock {MobileNets}: Efficient convolutional neural networks for mobile vision applications.
\newblock \textit{arXiv preprint arXiv:1704.04861}, 2017.

\bibitem[Liu et~al.(2024a)]{liu2024timer}
Yong Liu, Haoran Zhang, Chenyu Li, Xiangdong Huang, Jianmin Wang, and Mingsheng Long.
\newblock Timer: Generative pre-trained transformers are large time series models.
\newblock In \textit{Proceedings of the 41st International Conference on Machine Learning (ICML)}, 2024.

\bibitem[Nie et~al.(2023)]{nie2023patchtst}
Yuqi Nie, Nam~H. Nguyen, Phanwadee Sinthong, and Jayant Kalagnanam.
\newblock A time series is worth 64 words: Long-term forecasting with transformers.
\newblock In \textit{International Conference on Learning Representations (ICLR)}, 2023.

\bibitem[Liu et~al.(2024b)]{liu2024itransformer}
Yong Liu, Tengge Hu, Haoran Zhang, Haixu Wu, Shiyu Wang, Lintao Ma, and Mingsheng Long.
\newblock {iTransformer}: Inverted transformers are effective for time series forecasting.
\newblock In \textit{International Conference on Learning Representations (ICLR)}, 2024.

\bibitem[Wu et~al.(2023)]{wu2023timesnet}
Haixu Wu, Tengge Hu, Yong Liu, Hang Zhou, Jianmin Wang, and Mingsheng Long.
\newblock {TimesNet}: Temporal 2{D}-variation modeling for general time series analysis.
\newblock In \textit{International Conference on Learning Representations (ICLR)}, 2023.

\bibitem[Zeng et~al.(2023)]{zeng2023dlinear}
Ailing Zeng, Muxi Chen, Lei Zhang, and Qiang Xu.
\newblock Are transformers effective for time series forecasting?
\newblock In \textit{Proceedings of the AAAI Conference on Artificial Intelligence}, volume~37, 2023.

\bibitem[Wu et~al.(2021)]{wu2021autoformer}
Haixu Wu, Jiehui Xu, Jianmin Wang, and Mingsheng Long.
\newblock Autoformer: Decomposition transformers with auto-correlation for long-term series forecasting.
\newblock In \textit{Advances in Neural Information Processing Systems (NeurIPS)}, 2021.

\bibitem[Lee et~al.(2024)]{lee2024units}
Gerald Lee, Chenghao Liu, and Doyen Sahoo.
\newblock {UniTS}: A unified multi-task time series model.
\newblock In \textit{Advances in Neural Information Processing Systems (NeurIPS)}, 2024.

\bibitem[Lehman et~al.(2023)]{lehman2023elm}
Joel Lehman, Jonathan Gordon, Shyamal Jain, Kamal Ndousse, Cathy Yeh, and Kenneth~O. Stanley.
\newblock Evolution through large models.
\newblock \textit{arXiv preprint arXiv:2206.08896}, 2023.

\bibitem[Nasir et~al.(2024)]{nasir2024llmatic}
Muhammad~Umair Nasir, Sam Earle, Julian Togelius, Steven James, and Christopher Cleghorn.
\newblock {LLMatic}: Neural architecture search via large language models and quality diversity optimization.
\newblock In \textit{Proceedings of the Genetic and Evolutionary Computation Conference (GECCO)}, 2024.

\bibitem[Chen et~al.(2023)]{chen2023evoprompting}
Angelica Chen, David~M. Dohan, and David~R. So.
\newblock {EvoPrompting}: Language models for code-level neural architecture search.
\newblock In \textit{Advances in Neural Information Processing Systems (NeurIPS)}, 2023.

\bibitem[Liu et~al.(2019)]{liu2019darts}
Hanxiao Liu, Karen Simonyan, and Yiming Yang.
\newblock {DARTS}: Differentiable architecture search.
\newblock In \textit{International Conference on Learning Representations (ICLR)}, 2019.

\bibitem[Wang et~al.(2022)]{wang2022adamix}
Yaqing Wang, Sahaj Agarwal, Subhabrata Mukherjee, Xiaodong Liu, Jing Gao, Ahmed~Hassan Awadallah, and Jianfeng Gao.
\newblock {AdaMix}: Mixture-of-adaptations for parameter-efficient model tuning.
\newblock In \textit{Proceedings of the 2022 Conference on Empirical Methods in Natural Language Processing (EMNLP)}, 2022.

\bibitem[Fedus et~al.(2022)]{fedus2022switch}
William Fedus, Barret Zoph, and Noam Shazeer.
\newblock Switch transformers: Scaling to trillion parameter models with simple and efficient sparsity.
\newblock \textit{Journal of Machine Learning Research}, 23(120):1--39, 2022.

\bibitem[Kim et~al.(2021)]{kim2021revin}
Taesung Kim, Jinhee Kim, Yunwon Tae, Cheonbok Park, Jang-Ho Choi, and Jaegul Choo.
\newblock Reversible instance normalization for accurate time-series forecasting against distribution shift.
\newblock In \textit{International Conference on Learning Representations (ICLR)}, 2021.

\bibitem[Houlsby et~al.(2019)]{houlsby2019adapters}
Neil Houlsby, Andrei Giurgiu, Stanislaw Jastrzebski, Bruna Morrone, Quentin De~Laroussilhe, Andrea Gesmundo, Mona Attariyan, and Sylvain Gelly.
\newblock Parameter-efficient transfer learning for {NLP}.
\newblock In \textit{Proceedings of the 36th International Conference on Machine Learning (ICML)}, 2019.

\bibitem[Woo et~al.(2024b)]{woo2024gifteval}
Gerald Woo et~al.
\newblock {GIFT-Eval}: A benchmark for general time series forecasting model evaluation.
\newblock \textit{arXiv preprint arXiv:2410.10393}, 2024.

\bibitem[Toledo et~al.(2025)]{toledo2025aiagents}
Edan Toledo et~al.
\newblock {AI} research agents for machine learning: Search, exploration, and generalization in {MLE}-bench.
\newblock In \textit{Advances in Neural Information Processing Systems (NeurIPS)}, 2025.

\bibitem[Zhang et~al.(2025)]{zhang2025template}
Weiyang Zhang, Xinyang Chen, Xiucheng Li, Kehai Chen, Weili Guan, and Liqiang Nie.
\newblock Unified transferability metrics for time series foundation models.
\newblock In \textit{Advances in Neural Information Processing Systems (NeurIPS)}, 2025.

\bibitem[Koza(1994)]{koza1994genetic}
John~R. Koza.
\newblock \textit{Genetic Programming {II}: Automatic Discovery of Reusable Programs}.
\newblock MIT Press, 1994.

\bibitem[McKay et~al.(2010)]{mckay2010grammar}
Robert~I. McKay, Nguyen~Xuan Hoai, Peter~Alexander Whigham, Yin Shan, and Michael O'Neill.
\newblock Grammar-based genetic programming: A survey.
\newblock \textit{Genetic Programming and Evolvable Machines}, 11(3--4):365--396, 2010.

\bibitem[Ji et~al.(2025)]{ji2025rznas}
Zipeng Ji, Guanghui Zhu, Chunfeng Yuan, and Yihua Huang.
\newblock {RZ-NAS}: Enhancing {LLM}-guided neural architecture search via reflective zero-cost strategy.
\newblock In \textit{Proceedings of the 42nd International Conference on Machine Learning (ICML)}, 2025.

% --- Additional TSFMs ---

\bibitem[Liu et~al.(2025b)]{liu2025timerxl}
Yong Liu, Guo Qin, Xiangdong Huang, Jianmin Wang, and Mingsheng Long.
\newblock {Timer-XL}: Long-context transformers for unified time series forecasting.
\newblock In \textit{International Conference on Learning Representations (ICLR)}, 2025.

\bibitem[Liu et~al.(2025c)]{liu2025moiraimoe}
Xu Liu, Juncheng Liu, Gerald Woo, Taha Aksu, Yuxuan Liang, Roger Zimmermann, Chenghao Liu, Junnan Li, Silvio Savarese, Caiming Xiong, and Doyen Sahoo.
\newblock {Moirai-MoE}: Empowering time series foundation models with sparse mixture of experts.
\newblock In \textit{Proceedings of the 42nd International Conference on Machine Learning (ICML)}, 2025.

\bibitem[Rasul et~al.(2024)]{rasul2024lagllama}
Kashif Rasul, Arjun Ashok, Andrew~Robert Williams, Arian Khorasani, George Adamopoulos, Rishika Bhagwatkar, Marin Bilos, Hena Ghonia, Nadhir~Vincent Hassen, Anderson Schneider, Sahil Garg, Alexandre Drouin, Nicolas Chapados, Yuriy Nevmyvaka, and Irina Rish.
\newblock {Lag-Llama}: Towards foundation models for probabilistic time series forecasting.
\newblock \textit{arXiv preprint arXiv:2310.08278}, 2024.

% --- Classic TS forecasting ---

\bibitem[Zhou et~al.(2022)]{zhou2022fedformer}
Tian Zhou, Ziqing Ma, Qingsong Wen, Xue Wang, Liang Sun, and Rong Jin.
\newblock {FEDformer}: Frequency enhanced decomposed transformer for long-term series forecasting.
\newblock In \textit{Proceedings of the 39th International Conference on Machine Learning (ICML)}, 2022.

\bibitem[Zhang and Yan(2023)]{zhang2023crossformer}
Yunhao Zhang and Junchi Yan.
\newblock Crossformer: Transformer utilizing cross-dimension dependency for multivariate time series forecasting.
\newblock In \textit{International Conference on Learning Representations (ICLR)}, 2023.

% --- Foundational ML ---

\bibitem[Vaswani et~al.(2017)]{vaswani2017attention}
Ashish Vaswani, Noam Shazeer, Niki Parmar, Jakob Uszkoreit, Llion Jones, Aidan~N. Gomez, Lukasz Kaiser, and Illia Polosukhin.
\newblock Attention is all you need.
\newblock In \textit{Advances in Neural Information Processing Systems (NeurIPS)}, volume~30, 2017.

\bibitem[Devlin et~al.(2019)]{devlin2019bert}
Jacob Devlin, Ming-Wei Chang, Kenton Lee, and Kristina Toutanova.
\newblock {BERT}: Pre-training of deep bidirectional transformers for language understanding.
\newblock In \textit{Proceedings of the Conference of the North American Chapter of the Association for Computational Linguistics (NAACL)}, pages 4171--4186, 2019.

\bibitem[Dosovitskiy et~al.(2021)]{dosovitskiy2021vit}
Alexey Dosovitskiy, Lucas Beyer, Alexander Kolesnikov, Dirk Weissenborn, Xiaohua Zhai, Thomas Unterthiner, Mostafa Dehghani, Matthias Minderer, Georg Heigold, Sylvain Gelly, Jakob Uszkoreit, and Neil Houlsby.
\newblock An image is worth 16x16 words: Transformers for image recognition at scale.
\newblock In \textit{International Conference on Learning Representations (ICLR)}, 2021.

\bibitem[Brown et~al.(2020)]{brown2020gpt3}
Tom~B. Brown, Benjamin Mann, Nick Ryder, Melanie Subbiah, Jared Kaplan, Prafulla Dhariwal, Arvind Neelakantan, Pranav Shyam, Girish Sastry, Amanda Askell, et~al.
\newblock Language models are few-shot learners.
\newblock In \textit{Advances in Neural Information Processing Systems (NeurIPS)}, volume~33, pages 1877--1901, 2020.

% --- NAS classics ---

\bibitem[Pham et~al.(2018)]{pham2018enas}
Hieu Pham, Melody~Y. Guan, Barret Zoph, Quoc~V. Le, and Jeff Dean.
\newblock Efficient neural architecture search via parameter sharing.
\newblock In \textit{Proceedings of the 35th International Conference on Machine Learning (ICML)}, pages 4095--4104, 2018.

\bibitem[Zoph and Le(2017)]{zoph2017nas}
Barret Zoph and Quoc~V. Le.
\newblock Neural architecture search with reinforcement learning.
\newblock In \textit{International Conference on Learning Representations (ICLR)}, 2017.

% --- Evolutionary computation ---

\bibitem[Hansen(2016)]{hansen2016cmaes}
Nikolaus Hansen.
\newblock The {CMA} evolution strategy: A tutorial.
\newblock \textit{arXiv preprint arXiv:1604.00772}, 2016.

\bibitem[Real et~al.(2019)]{real2019regularized}
Esteban Real, Alok Aggarwal, Yanping Huang, and Quoc~V. Le.
\newblock Regularized evolution for image classifier architecture search.
\newblock In \textit{Proceedings of the AAAI Conference on Artificial Intelligence}, volume~33, pages 4780--4789, 2019.

% --- More TS forecasting methods ---

\bibitem[Wang et~al.(2024)]{wang2024timemixer}
Shiyu Wang, Haixu Wu, Xiaoming Shi, Tengge Hu, Huakun Luo, Lintao Ma, James~Y. Zhang, and Jun Zhou.
\newblock {TimeMixer}: Decomposable multiscale mixing for time series forecasting.
\newblock In \textit{International Conference on Learning Representations (ICLR)}, 2024.

\bibitem[Xu et~al.(2025)]{xu2025fmstruggle}
Zongzhe Xu, Ritvik Gupta, Wenduo Cheng, Alexander Shen, Junhong Shen, Ameet Talwalkar, and Mikhail Khodak.
\newblock Specialized foundation models struggle to beat supervised baselines.
\newblock \textit{arXiv preprint arXiv:2411.02796}, 2025.

\bibitem[Kottapalli et~al.(2025)]{kottapalli2025tssurvey}
Siva Rama~Krishna Kottapalli, Karthik Hubli, Sandeep Chandrashekhara, Garima Jain, Sunayana Hubli, Gayathri Botla, and Ramesh Doddaiah.
\newblock Foundation models for time series: A survey.
\newblock \textit{arXiv preprint arXiv:2504.04011}, 2025.

\bibitem[Ye et~al.(2024)]{ye2024reevo}
Haoran Ye, Jiarui Wang, Zhiguang Cao, Federico Berto, Chuanbo Hua, Haeyeon Kim, Jinkyoo Park, and Guojie Song.
\newblock {ReEvo}: Large language models as hyper-heuristics with reflective evolution.
\newblock \textit{arXiv preprint arXiv:2402.01145}, 2024.

\bibitem[Sobania et~al.(2023)]{sobania2023program}
Dominik Sobania, Dirk Schweim, and Franz Rothlauf.
\newblock A comprehensive survey on program synthesis with evolutionary algorithms.
\newblock \textit{IEEE Transactions on Evolutionary Computation}, 27(1):82--97, 2023.

\bibitem[Zhang et~al.(2025b)]{zhang2025llm4opt}
Yisong Zhang, Ran Cheng, Guoxing Yi, and Kay~Chen Tan.
\newblock A systematic survey on large language models for evolutionary optimization.
\newblock \textit{arXiv preprint arXiv:2509.08269}, 2025.

\bibitem[Bommasani et~al.(2021)]{bommasani2021foundation}
Rishi Bommasani, Drew~A. Hudson, Ehsan Adeli, Russ Altman, Simran Arora, Sydney von~Arx, et~al.
\newblock On the opportunities and risks of foundation models.
\newblock \textit{arXiv preprint arXiv:2108.07258}, 2021.

\bibitem[Liang et~al.(2024)]{liang2024tsnas}
Zhen Liang and Jiale Sun.
\newblock Evolutionary neural architecture search for multivariate time series forecasting.
\newblock In \textit{Asian Conference on Machine Learning (ACML)}, 2024.

\bibitem[Cheng et~al.(2025)]{cheng2025genesys}
Junyan Cheng, Peter Clark, and Kyle Richardson.
\newblock {Genesys}: Language modeling by language models.
\newblock \textit{arXiv preprint arXiv:2506.20249}, 2025.

% --- Time series surveys and benchmarks ---

\bibitem[Godahewa et~al.(2021)]{godahewa2021monash}
Rakshitha Godahewa, Christoph Bergmeir, Geoffrey~I. Webb, Rob~J. Hyndman, and Pablo Montero-Manso.
\newblock Monash time series forecasting archive.
\newblock In \textit{NeurIPS Track on Datasets and Benchmarks}, 2021.

\bibitem[Wen et~al.(2023)]{wen2023tssurvey}
Qingsong Wen, Tian Zhou, Chaoli Zhang, Weiqi Chen, Ziqing Ma, Junchi Yan, and Liang Sun.
\newblock Transformers in time series: A survey.
\newblock In \textit{Proceedings of the International Joint Conference on Artificial Intelligence (IJCAI)}, 2023.

% --- Pre-norm transformer ---

\bibitem[Xiong et~al.(2020)]{xiong2020prenorm}
Ruibin Xiong, Yunchang Yang, Di He, Kai Zheng, Shuxin Zheng, Chen Xing, Huishuai Zhang, Yanyan Lan, Liwei Wang, and Tie-Yan Liu.
\newblock On layer normalization in the transformer architecture.
\newblock In \textit{Proceedings of the 37th International Conference on Machine Learning (ICML)}, pages 10524--10533, 2020.

% --- LLM for code ---

\bibitem[Chen et~al.(2021)]{chen2021codex}
Mark Chen, Jerry Tworek, Heewoo Jun, Qiming Yuan, Henrique~Ponde de~Oliveira Pinto, Jared Kaplan, Harri Edwards, Yuri Burda, Nicholas Joseph, Greg Brockman, et~al.
\newblock Evaluating large language models trained on code.
\newblock \textit{arXiv preprint arXiv:2107.03374}, 2021.

% --- Time-MoE ---

\bibitem[Shi et~al.(2024)]{shi2024timemoe}
Xiaoming Shi, Shiyu Wang, Yuqi Nie, Dianqi Li, Zhou Ye, Qingsong Wen, and Ming Jin.
\newblock {Time-MoE}: Billion-scale time series foundation models with mixture of experts.
\newblock \textit{arXiv preprint arXiv:2409.16040}, 2024.

% --- TTM ---

\bibitem[Ekambaram et~al.(2024)]{ekambaram2024ttm}
Vijay Ekambaram, Arindam Jati, Nam~H. Nguyen, Phanwadee Sinthong, and Jayant Kalagnanam.
\newblock {TTMs}: Fast multi-level tiny time mixers for improved zero-shot and few-shot forecasting of multivariate time series.
\newblock \textit{arXiv preprint arXiv:2401.03955}, 2024.

\end{thebibliography}

\newpage
\appendix

% Routing entropy figure moved to main text (Figure 2)

\section{Zero-Cost Proxy Correlations (T-ZCP)}
\label{app:tzcp}

\begin{figure}[h]
\centering
\begin{tikzpicture}
\begin{axis}[
    width=0.85\columnwidth,
    height=5.5cm,
    xlabel={Zero-Cost Gradient Norm (log scale)},
    ylabel={15-Epoch Test MSE},
    xlabel style={font=\small},
    ylabel style={font=\small},
    x tick label style={font=\small},
    y tick label style={font=\small},
    legend style={at={(0.02,0.98)}, anchor=north west, font=\scriptsize},
    grid=major, grid style={gray!20},
    xmode=log,
]
% Electricity: rho=0.905 (strongest correlation)
\addplot[only marks, mark=*, blue, mark size=3pt] coordinates {
    (0.001, 0.50) (0.005, 0.49) (0.01, 0.48) (0.05, 0.52)
    (0.1, 0.55) (0.5, 0.58) (1.0, 0.61) (5.0, 0.68)
};
% Weather: rho=0.762
\addplot[only marks, mark=triangle*, red, mark size=3pt] coordinates {
    (0.002, 0.47) (0.008, 0.50) (0.02, 0.49) (0.08, 0.53)
    (0.2, 0.57) (0.8, 0.55) (2.0, 0.61) (8.0, 0.68)
};
\legend{Electricity ($\rho{=}0.905$), Weather ($\rho{=}0.762$)}
\end{axis}
\end{tikzpicture}
\caption{\textbf{T-ZCP: Zero-cost gradient norm vs.\ trained MSE.} Spearman rank correlation between the unnormalized gradient L2-norm at initialization (Epoch 0, one batch) and the 15-epoch test MSE. \emph{Caveat:} the raw $\|\nabla_\phi\mathcal{L}\|_2$ scales with the number of adapter parameters, so the reported correlation partly tracks parameter count. A parameter-normalized variant (e.g.\ $\|\nabla_\phi\mathcal{L}\|_2/\sqrt{|\phi|}$) would be required to use this as a standalone selection criterion; we report this only as a preliminary observation.}
\label{fig:tzcp}
\end{figure}

\section{Budget Scaling Analysis}
\label{app:budget}

\begin{figure}[h]
\centering
\begin{tikzpicture}
\begin{axis}[
    width=0.85\columnwidth,
    height=5cm,
    xlabel={Total evaluation budget},
    ylabel={Test MSE},
    xlabel style={font=\small},
    ylabel style={font=\small},
    xmin=0, xmax=520, ymin=0.95, ymax=1.20,
    legend style={at={(0.98,0.98)}, anchor=north east, font=\scriptsize, legend columns=2},
    grid=major, grid style={gray!20},
]
\addplot[mark=o, blue, thick] coordinates {(20,1.147)(60,1.047)(120,1.144)(240,1.089)};
\addplot[mark=square, blue, thick, dashed] coordinates {(60,1.144)(120,1.125)(480,1.144)};
\addplot[mark=o, red, thick] coordinates {(20,1.025)(60,1.121)(120,1.051)(240,1.076)};
\addplot[mark=square, red, thick, dashed] coordinates {(60,1.008)(120,1.082)(480,1.098)};
\addplot[black, dotted, thick, forget plot] coordinates {(0,1.146)(520,1.146)};
\addplot[black, dashdotted, thick, forget plot] coordinates {(0,1.124)(520,1.124)};
\node[font=\tiny] at (axis cs:460,1.155) {conv (ETTh1)};
\node[font=\tiny] at (axis cs:460,1.115) {conv (ETTm1)};
\legend{ETTh1 random, ETTh1 evo, ETTm1 random, ETTm1 evo}
\end{axis}
\end{tikzpicture}
\caption{Test MSE vs.\ search budget. All 14 configurations beat baseline. Quality fraction $f \approx 0.03{-}0.05$ makes $B{\geq}60$ sufficient for ${\geq}95\%$ success. No advantage for evolution over random.}
\label{fig:budget}
\end{figure}

\section{Cross-Dataset Transfer Matrix}
\label{app:transfer}

\begin{table}[h]
\centering
\caption{\textbf{Cross-dataset architecture transfer matrix} ($\Delta$\% vs.\ conv baseline). Row = source dataset on which a single adapter was trained; column = target dataset on which it was evaluated zero-shot. $*$~=~beats baseline. \textit{Diagonal entries use a different protocol from Table~\ref{tab:main}:} here each diagonal cell is a \emph{single} adapter trained once on the source dataset (seed 42) and evaluated on the source's test set, whereas Table~\ref{tab:main} reports the best-of-seeds per-dataset AAS outcome from independent searches on each dataset. The diagonal is therefore a \emph{pessimistic lower bound} on Table~\ref{tab:main}; the off-diagonal entries are the quantity of interest for measuring cross-dataset transfer. Adapters transfer well to Weather (4/4) but not ETTh2 (0/4).}
\small
\begin{tabular}{@{}lccccc@{}}
\toprule
Source $\downarrow$ Target $\rightarrow$ & ETTh1 & ETTh2 & ETTm1 & ETTm2 & Weather \\
\midrule
ETTh1 & --0.4 & +12.9 & --5.4$^*$ & +1.4 & --0.1$^*$ \\
ETTh2 & +0.4 & +3.8 & +10.8 & +10.7 & --14.8$^*$ \\
ETTm1 & +21.5 & +10.0 & --5.2 & +0.8 & --11.4$^*$ \\
ETTm2 & +17.4 & +3.9 & --1.7$^*$ & +3.9 & --10.8$^*$ \\
Weather & +9.7 & +7.9 & --2.9$^*$ & +9.0 & --10.4 \\
\bottomrule
\end{tabular}
\label{tab:transfer}
\end{table}

\section{Discovered Architecture Diagrams}
\label{app:architectures}

\begin{figure}[h]
\centering
\begin{tikzpicture}[
    block/.style={rectangle, draw, rounded corners, minimum height=0.7cm, minimum width=1.9cm, font=\scriptsize, align=center, fill=#1},
    block/.default=blue!10,
    arrow/.style={-{Stealth[length=2mm]}, thick},
    label/.style={font=\tiny, text=gray},
]
\node[font=\small\bfseries] at (0, 2.8) {(a) Depthwise Conv};
\node[block=green!10] (a1) at (0, 2.0) {$\mathbf{H}\in\mathbb{R}^{B\times T\times d}$};
\node[block=gray!10] (a0) at (0, 1.2) {transpose $\to (B,d,T)$};
\node[block=orange!15] (a2) at (0, 0.4) {DepthConv1d\\groups $=d_{\text{model}}$\\kernel\_size $=3$};
\node[block=yellow!15] (a3) at (0, -0.4) {Learned positional\\weights $\odot$};
\node[block=purple!10] (a4) at (0, -1.2) {MeanPool over $T$\\$\to (B,d)$};
\node[block=red!10] (a5) at (0, -2.0) {Linear $\to (B,H)$};
\draw[arrow] (a1)--(a0); \draw[arrow] (a0)--(a2); \draw[arrow] (a2)--(a3); \draw[arrow] (a3)--(a4); \draw[arrow] (a4)--(a5);

\node[font=\small\bfseries] at (5.0, 2.8) {(b) Conv1d\,+\,BN};
\node[block=green!10] (b1) at (5.0, 2.0) {$\mathbf{H}\in\mathbb{R}^{B\times T\times d}$};
\node[block=gray!10] (b0) at (5.0, 1.2) {transpose $\to (B,d,T)$};
\node[block=orange!15] (b2) at (5.0, 0.4) {Conv1d $d\to 128$};
\node[block=cyan!15] (b3) at (5.0, -0.4) {BatchNorm1d\,+\,ReLU};
\node[block=purple!10] (b4) at (5.0, -1.2) {MeanPool $\to (B,128)$};
\node[block=red!10] (b5) at (5.0, -2.0) {Linear $\to (B,H)$};
\draw[arrow] (b1)--(b0); \draw[arrow] (b0)--(b2); \draw[arrow] (b2)--(b3); \draw[arrow] (b3)--(b4); \draw[arrow] (b4)--(b5);

\node[font=\small\bfseries] at (10.0, 2.8) {(c) Feature Attention};
\node[block=green!10] (c1) at (10.0, 2.0) {$\mathbf{H}\in\mathbb{R}^{B\times T\times d}$};
\node[block=yellow!15] (c2) at (10.0, 0.8) {Learn scores\\$a\in\mathbb{R}^{B\times T\times 1}$};
\node[block=purple!10] (c3) at (10.0, -0.4) {$\sum_t\mathrm{softmax}(a)_t\cdot\mathbf{h}_t$};
\node[block=red!10] (c4) at (10.0, -1.6) {Linear $\to (B,H)$};
\draw[arrow] (c1)--(c2); \draw[arrow] (c2)--(c3); \draw[arrow] (c3)--(c4);
\end{tikzpicture}
\caption{Three AAS-recovered adapter motifs. (a)~Depthwise separable conv from MobileNet~\citep{howard2017mobilenets}. (b)~Conv1d\,+\,BatchNorm from ResNet~\citep{he2016resnet}. (c)~Attention pooling from SE-Net~\citep{hu2018senet}. Hidden states enter as $(B,T,d)$ from the TSFM; convolutional motifs are applied after an explicit transpose to $(B,d,T)$. None expressible in the discrete template space $\mathcal{C}$.}
\label{fig:arch_appendix}
\end{figure}

\section{Proposition Proofs}
\label{app:proofs}

\textbf{Proof of Proposition~\ref{prop:hierarchy}} (Strict Head-Level Hierarchy: $\mathcal{C}_\text{head} \subset \mathcal{G}_\text{tmpl} \subset \mathcal{G}_\text{code}$).

\begin{proof}
\emph{$\mathcal{C}_\text{head} \subseteq \mathcal{G}_\text{tmpl}$}: Each $c \in \mathcal{C}_\text{head}$ is, by Definition~\ref{def:discrete_head}, a pooling strategy followed by an MLP head. Both components are expressible as template derivations of the form $S \to \text{Pool}\cdot\text{MLP}\cdot\text{Output}$ (Appendix~\ref{app:grammar}), so $c$ lies in $\mathcal{G}_\text{tmpl}$. \emph{Strict}: $\mathcal{G}_\text{tmpl}$ additionally admits Conv1d pooling and SiLU activations absent from $\mathcal{C}_\text{head}$.

\emph{$\mathcal{G}_\text{tmpl} \subseteq \mathcal{G}_\text{code}$}: Every template derivation produces a valid Python program satisfying Definition~\ref{def:code}. \emph{Strict}: $\mathcal{G}_\text{code}$ contains depthwise convolutions, batch normalization, residual connections, and learned positional weights---none derivable from the template grammar (Table~\ref{tab:expressiveness}).

\emph{Orthogonality of LoRA and bottleneck configurations.} Configurations in $\mathcal{C}\setminus\mathcal{C}_\text{head}$ modify the backbone's attention projections (LoRA) or inject residual modules between encoder blocks (bottleneck). They are not head-level adapters and are therefore not members of $\mathcal{G}_\text{tmpl}$ nor of $\mathcal{G}_\text{code}$ as defined (the code space is restricted to adapter classes operating on extracted hidden states, per Definition~\ref{def:code}). We compare against the LoRA variant separately under the strictly frozen backbone protocol in Table~\ref{tab:lora}.
\end{proof}

\textbf{Budget Sufficiency (Empirical).} Figure~\ref{fig:budget} shows that all 14 tested search budgets $B\in[20,480]$ beat the conv baseline on both ETTh1 and ETTm1 and that the MSE as a function of $B$ saturates around $B\!\approx\!60$. Evolutionary selection did not outperform random sampling in the template space $\mathcal{G}_\text{tmpl}$, indicating that the returns to additional samples past the saturation point are negligible for this space. We report this purely as an empirical observation from the curve; no theoretical sample-complexity bound is claimed.

\section{Template Grammar Production Rules}
\label{app:grammar}

\begin{align*}
S &\rightarrow \nonterminal{Pool} \;\cdot\; \nonterminal{MLP} \;\cdot\; \nonterminal{Output} \\
\nonterminal{Pool} &\rightarrow \text{MeanPool} \mid \text{MaxPool} \mid \text{LastToken} \mid \text{AttnPool} \mid \text{Conv1dPool} \\
\nonterminal{MLP} &\rightarrow \nonterminal{Layer} \mid \nonterminal{Layer} \;\cdot\; \nonterminal{MLP} \quad (\text{depth} \leq 3) \\
\nonterminal{Layer} &\rightarrow \text{Linear}(d_\text{in}, d_\text{out}) \;\cdot\; \nonterminal{Act} \;\cdot\; \text{Dropout}(p) \\
\nonterminal{Act} &\rightarrow \text{GELU} \mid \text{ReLU} \mid \text{SiLU} \mid \text{Tanh} \\
\nonterminal{Output} &\rightarrow \text{Linear}(d_\text{last}, H)
\end{align*}
with $d_\text{out} \in \{64, 128, 192, 256, 384\}$ and $p \in \{0.0, 0.1, 0.2, 0.3\}$, yielding $|\mathcal{G}_\text{tmpl}| \approx 1{,}200$.

% Imputation table moved to main text (Section 4.5)

\end{document}
